---
title: "EDA - all 4 samples per patient"
author: "Valerio Reffo"
date: "2026-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(spatstat)
library(tidyverse)
library(purrr)
library(fda)
library(readxl)
library(pals)
```


```{r}
dati = read_csv("dataset/CRC_clusters_neighborhoods_markers.csv")
```

We have 4 samples for each patient

```{r}
samples_list <- split(dati, dati$`File Name`)
length(samples_list)
```

Create a single ppp for each sample

```{r}
marks_name <- c("ClusterName", "neighborhood name")
ppp_pats <- map(samples_list, function(pat) {
  # Note: Ensure 'cell_type' column exists in your data. 
  # If it is named 'ClusterID' or something else, update it here.
  ppp_creation(pat, marks_name, "X:X", "Y:Y")
})

# Check the result
# ppp_pats is now a list of ppp objects, named by patient ID
print(names(ppp_pats))
```

Patients data

```{r}
pat_data <- read_excel("dataset/CRC_TMAs_patient_annotations.xlsx", n_max = 35)
pat_data <- pat_data |>  mutate(Patient = as.character(Patient))
head(pat_data)
```

Let's create an object that can accomodate every kind of object we would need to analyze the dataset,
for each sample we can add info about, the object is an Hyper Data frame

```{r}
ppp_hf <- as.hyperframe(ppp = as.solist(ppp_pats))
ppp_hf <- cbind(ppp_hf, sample_name = unique(dati$`File Name`))
```

Let's add all the info about the patients
```{r}
# 1. Extract unique metadata for each sample (File Name)
# We select the key and the columns you want to add
metadata <- dati |> 
  dplyr::select(`File Name`, 
    Region, 
    groups, 
    patients, 
    spots) |> 
  # Keep only one row per unique 'File Name'
  distinct(`File Name`, .keep_all = TRUE)

# 2. Align the metadata with the hyperframe order
# We match 'sample_name' from the hyperframe to 'File Name' in the metadata
matching_indices <- match(ppp_hf$sample_name, metadata$`File Name`)

# 3. Add the columns to the Hyperframe
# We assign all columns except the key ("File Name")
cols_to_add <- setdiff(colnames(metadata), "File Name")
ppp_hf[, cols_to_add] <- metadata[matching_indices, cols_to_add]

# Verify the result
print(ppp_hf)
```


```{r}
ppp_df <- dati |> 
  dplyr::select(`File Name`, 
    Region, 
    groups, 
    patients, 
    spots,
    ClusterName,
    `neighborhood name`,
    `X:X`,
    `Y:Y`
    ) |> 
  mutate(
    sample = `File Name`,
    patients = as.character(patients),
    x = `X:X`,
    y = `Y:Y`
  )
ppp_df <- left_join(ppp_df, pat_data, by = c("patients" = "Patient"))
```

Save the object to work with
```{r}
saveRDS(ppp_df, file = "dataset/data_for_spatialplot.rds")
saveRDS(ppp_hf, file = "dataset/hyperframe_140samples_integrated_info.rds")
```


# Plot

by patients, of course it's wrong
```{r}

ggplot(ppp_df, aes(x = x, y = y, color = ClusterName)) +
  # Use a small size and some transparency (alpha) because you have 250k points
  geom_point(size = 0.1, alpha = 0.5) +
  
  # This creates a separate plot for each patient
  facet_wrap(~ patients, scales = "free") +
  
  # Essential for spatial data: keeps the 1:1 aspect ratio of the tissue
  #coord_fixed() +
  
  # Clean up the look
  theme_minimal() +
  labs(title = "Spatial Distribution of Clusters by Patient",
       x = "X Coordinate",
       y = "Y Coordinate",
       color = "Cell Type") +
  
  # Optional: customize the legend if there are too many categories
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

```

by sample
```{r}
group_pats <- ppp_df |>
  distinct(patients) |>
  mutate(page = ntile(patients, 4)) # Divides patients into 4 equal-sized groups


all_plots <- ppp_df |> 
  left_join(group_pats, by = "patients") |> 
  group_split(page) |>
  map(\(page) {
    
    num_clusters <- length(unique(ppp_df$ClusterName))
    my_colors <- as.vector(pals::glasbey(num_clusters))
    # .x represents each chunk of data
    ggplot(page, aes(x = x, y = y, color = ClusterName)) +
      geom_point(size = 0.1, alpha = 0.6) +
      facet_wrap(~ sample, scales = "free", ncol = 6) +
      scale_color_manual(values = my_colors) + 
      #coord_fixed() +
      theme_minimal() +
      theme(
        legend.position = "left",             # Sposta a sinistra
        legend.text = element_text(size = 6),  # Testo piccolo per risparmiare spazio
        legend.title = element_text(size = 8), # Titolo piccolo
        legend.key.size = unit(0.2, "cm"),     # Rimpicciolisce i quadratini della legenda
        legend.spacing.x = unit(0.1, "cm")     # Riduce lo spazio tra simbolo e testo
      ) +
      guides(color = guide_legend(
        override.aes = list(size = 3, alpha = 1), # Simboli grandi nella legenda
        ncol = 1                                  # Una sola colonna per occupare meno larghezza
      )) +
      
      labs(title = paste("Patient Spatial Distribution"),
           x = "X Coordinate", y = "Y Coordinate")
  })

iwalk(all_plots, ~{
  ggsave(
    filename = paste0("Patient_Page_", .y, ".png"), 
    plot = .x, 
    width = 20,     # Dimensioni grandi per mantenere la risoluzione
    height = 15, 
    dpi = 300       # Alta qualità
  )
})# Now 'all_plots' is a list of 4 ggplot objects.
# You can view them:
all_plots[[1]]
all_plots[[2]]

# Or save them all to a PDF in one line
walk(all_plots, print) # 'walk' is like map but for actions (printing)
```

Reduce the number of clusters
```{r}
unique(dati$ClusterName)
```

"tumor cells"  

"granulocytes"               
"nerves"                     
"vasculature"                
"adipocytes"                 
"plasma cells"              

"CD4+ T cells CD45RO+"       
"CD4+ T cells GATA3+"        
"CD3+ T cells" 
"CD8+ T cells"               
"CD4+ T cells"              

"smooth muscle"             
"stroma"                     

"CD68+ macrophages"         
"CD68+ macrophages GzmB+"    
"CD68+CD163+ macrophages"    
"CD11b+CD68+ macrophages"    
"CD163+ macrophages"        

"Tregs"                      
"dirt"                       
"NK cells"                   
"lymphatics"                 

"immune cells"              

"B cells"                    

"undefined"                  

"tumor cells / immune cells" 
"immune cells / vasculature" 
"CD11c+ DCs"                 
"CD11b+ monocytes"           







