---
title: "eda_mendeley_data"
author: "Valerio Reffo"
date: "2026-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(readr)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(purrr)
library(spatstat)
library(fda)
library(forcats)
```

```{r}
ppp_pats <- readRDS("dataset/ppp_patients_data.rds")
```

patients info
```{r}
pat_data <- read_excel("dataset/CRC_TMAs_patient_annotations.xlsx", n_max = 35)
pat_data <- pat_data |>  mutate(Patient = as.character(Patient))
```


Subsetting only the 2 marks of interest, Tumor and Immune cells


```{r}
ppp_pats <- map(ppp_pats, \(pat) {subset(pat, marks %in% c("Tumor", "Immune-infiltrated stroma"), drop = TRUE)})
```

# Plots of the point processes

Let's create the df for the plots with annotations about patients 
```{r}
ppp_df <- imap_dfr(ppp_pats, function(ppp, name) {
  # Defensive: skip empty ppp objects
  if (ppp$n == 0) return(NULL)
  data.frame(
    x = ppp$x,
    y = ppp$y,
    mark = marks(ppp),
    patient = name
  )
}) 
ppp_df <- left_join(ppp_df, pat_data, by = c("patient" = "Patient"))
```



```{r}
ggplot(ppp_df, aes(x = x, y = y, color = mark)) +
  geom_point(size = 0.4, alpha = 0.7) +
  facet_wrap(~ patient) +
  coord_fixed() +
  labs(title = "Points from multiple patients", color = "Cell type") +
  theme_minimal()
```

Histogram of the 2 marks for each patient
```{r}
ggplot(ppp_df, aes(x = mark, fill = mark)) +
  geom_bar() +
  facet_wrap(~ patient, scales = "free_y") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    x = "Cell type (mark)",
    y = "Count",
    title = "Histogram of marks per patient"
  )
```

Explore the effect of single covariates, let's compare the number of points in the point processes for each variable

Let's work with the hyperframe so it's easier (?) to manage the variables
```{r}
pats_names <- sapply(names(ppp_pats), \(name) {paste0("Patient",name)}, simplify = T)
names(ppp_pats) <- pats_names
pats_hf <- as.solist(ppp_pats)
(pats_hf <- as.hyperframe(ppp = pats_hf, row.names = names(ppp_pats)))
pats_hf$n <- with(pats_hf, npoints(ppp))
pats_hf$patients <- names(ppp_pats)
pats_hf$area <- with(pats_hf, area.owin(ppp$window))
pats_hf <- cbind(pats_hf, pat_data[!(rownames(pat_data) %in% c(9,19) ),colnames(pat_data) %in% c("Group", "Sex", "Age", "Full_Histology", "Full_Tumor_Location",  "Simple_Tumor_Location")])
```


```{r}
var_colors <- c("Group", "Sex", "Full_Tumor_Location", "Full_Histology")

map(var_colors, \(color){
  form <- as.formula(paste("sqrt(n) ~", color))
  boxplot(form, data = pats_hf)
  v <- pretty(pats_hf$n); axis(4, at=sqrt(v), labels=v)
  title(paste0(color, " variable"))
})
```

With mean intensity
```{r}
aggregate(n/area ~ Group, data = pats_hf, FUN = mean)

map(var_colors, \(color){
  form <- as.formula(paste("n/area ~", color))
  aggregate(form, data = pats_hf, FUN = \(x) {mean(x)*1000})
})
```
IF we have windows of different area we better weight the average

Group
```{r}
ntot <- sapply(split(pats_hf$n, pats_hf$Group), sum)
atot <- sapply(split(pats_hf$area, pats_hf$Group), sum)
ntot/atot*1000
```

Sex
```{r}
ntot <- sapply(split(pats_hf$n, pats_hf$Sex), sum)
atot <- sapply(split(pats_hf$area, pats_hf$Sex), sum)
ntot/atot*1000
```

Full Tumor Location
```{r}
ntot <- sapply(split(pats_hf$n, pats_hf$Full_Tumor_Location), sum)
atot <- sapply(split(pats_hf$area, pats_hf$Full_Tumor_Location), sum)
ntot/atot*1000
```

Full Histology
```{r}
ntot <- sapply(split(pats_hf$n, pats_hf$Full_Histology), sum)
atot <- sapply(split(pats_hf$area, pats_hf$Full_Histology), sum)
ntot/atot*1000
```


Fun for plotting different categorical variables
```{r}
fsplit_plot <- function(data, factor) {
  factor_quo <- enquo(factor)
  split_list <- data |> group_split(!!factor_quo)
  group_label <- as_label(factor_quo)
  
  walk(split_list, function(single_group) {
    # Extract the group value for this group (as character)
    gval <- unique(single_group[[group_label]])[1]
    plt <- ggplot(single_group, aes(x = x, y = y, color = mark)) +
      geom_point(size = 0.7, alpha = 0.7) +
      facet_wrap(~ patient) +
      coord_fixed() +
      labs(title = paste("Points from patients -", group_label, "=", gval),
           color = "Cell type") +
      theme_minimal()
    print(plt)
  })
}
```

## split by categories

Split by "Group":
                  Group == 1 : Crohn's like reaction (CLR)
                  Group == 2 : diffuse inflammatory infiltration (DII)

```{r}
fsplit_plot(ppp_df, Group)
```

Sex
```{r}
fsplit_plot(ppp_df, Sex)
```

Full histology

```{r}
fsplit_plot(ppp_df, Full_Histology)
```

Full_Tumor_Location

```{r}
fsplit_plot(ppp_df, Full_Tumor_Location)
```

```{r}
Lfun_list <- Lcross_smooth(ppp_pats, centrata = T, rmax = 150)
```

Plot of the Lcross smoothed function colored by different patients' characteristics

```{r}
var_colors <- c("Group", "Sex", "Full_Tumor_Location", "Full_Histology")

walk(var_colors, \(color) {
  plt <- plot_fd_gg_meta(
    fdobj    = Lfun_list$fd,
    meta_df  = pat_data,
    id_col   = "Patient",
    color_by = color,
    facet_by = NULL,
    label_curves = TRUE,    # Add the labels
    label_var = "Patient",   # The variable to show as text,
    grid_length = 50
  )
  print(plt)
})

```

Facet by Sex
```{r}
var_colors <- c("Group", "Full_Tumor_Location", "Full_Histology")

walk(var_colors, \(color) {
  plt <- plot_fd_gg_meta(
    fdobj    = Lfun_list$fd,
    meta_df  = pat_data,
    id_col   = "Patient",
    color_by = color,
    facet_by = "Sex",
    label_curves = TRUE,    # Add the labels
    label_var = "Patient",   # The variable to show as text,
    grid_length = 50
  )
  print(plt)
})
```

Facet by Group
```{r}
var_colors <- c("Sex", "Full_Tumor_Location", "Full_Histology")

walk(var_colors, \(color) {
  plt <- plot_fd_gg_meta(
    fdobj    = Lfun_list$fd,
    meta_df  = pat_data,
    id_col   = "Patient",
    color_by = color,
    facet_by = "Group",
    label_curves = TRUE,    # Add the labels
    label_var = "Patient",   # The variable to show as text,
    grid_length = 50
  )
  print(plt)
})
```

Facet by Full_Histology
```{r}
var_colors <- c("Group", "Sex", "Full_Tumor_Location")

walk(var_colors, \(color) {
  plt <- plot_fd_gg_meta(
    fdobj    = Lfun_list$fd,
    meta_df  = pat_data,
    id_col   = "Patient",
    color_by = color,
    facet_by = "Full_Histology",
    label_curves = TRUE,    # Add the labels
    label_var = "Patient",   # The variable to show as text,
    grid_length = 50
  )
  print(plt)
})
```

Facet by Full_Tumor_Location
```{r}
var_colors <- c("Group", "Sex", "Full_Histology")

walk(var_colors, \(color) {
  plt <- plot_fd_gg_meta(
    fdobj    = Lfun_list$fd,
    meta_df  = pat_data,
    id_col   = "Patient",
    color_by = color,
    facet_by = "Full_Tumor_Location",
    label_curves = TRUE,    # Add the labels
    label_var = "Patient",   # The variable to show as text,
    grid_length = 50
  )
  print(plt)
})
```


# spatial plot group by Lcross behaviour

Function to plot a subset of the ppp
```{r}
multi_patient_plot <- function(data, patients, filter_factor = NULL, factor_name = NULL, pp_type) {
  # Coerce patient types to character for robust comparison
  data$patient <- as.character(data$patient)
  patients <- as.character(patients)
  plot_df <- data %>% filter(patient %in% patients)
  
  # Optionally filter by group/factor
  if (!is.null(filter_factor) && !is.null(factor_name)) {
    plot_df <- plot_df %>% filter(.data[[factor_name]] == filter_factor)
    main_title <- paste0("Patients: ", paste(patients, collapse=", "), 
                         " (", factor_name, "=", filter_factor, ")")
  } else {
    main_title <- paste0("Patients: ", paste(patients, collapse=", "), ". Type: ", pp_type)
  }

  ggplot(plot_df, aes(x = x, y = y, color = mark)) +
    geom_point(size = 0.4, alpha = 0.7) +
    facet_wrap(~ patient) +
    coord_fixed() +
    labs(title = main_title, color = "Cell type") +
    theme_minimal()
}
```

Patients in which the 2 cell type show the strongest association then in the other patients
```{r}
clust_plot <- multi_patient_plot(ppp_df, patients = c(29, 11, 6), pp_type = "Positive association")
```

Patients which show negative association between the 2 cell types

```{r}
reg_plot <- multi_patient_plot(ppp_df, patients = c(14, 13, 5), pp_type = "Negative association")
```

At last the patients showing indipendence between the cell types
```{r}
csr_plot <- multi_patient_plot(ppp_df, patients = c(2, 23, 16), pp_type = "Indipendence")
```

```{r}
clust_plot/csr_plot/reg_plot
```

