---
title: "Esplorazione Mendeley dataset"
author: "Valerio Reffo"
date: "2026-01-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(spatstat)
library(tidyverse)
library(purrr)
```


```{r}
dat = read_csv("dataset/CRC_clusters_neighborhoods_markers.csv")
pat1 <- dat[dat$patients == 1,]
pat1 |> ggplot(aes(`X:X`, `Y:Y`, color = `neighborhood name`)) + geom_point()
```

```{r}
table(dat$patients)
```

```{r}
table(dat$Region, dat$patients)
```

Ci sono 2 regioni per ogni paziente

```{r}
pat1[pat1$Region=="reg001",] |> 
  ggplot(aes(`X:X`, `Y:Y`, color = `neighborhood name`)) + geom_point()
```

```{r}
pat1[pat1$Region=="reg002",] |> 
  ggplot(aes(`X:X`, `Y:Y`, color = `neighborhood name`)) + geom_point()
```

# Pulizia e sistemazione dataset

Inserisco una variabile per identificare le due regioni per ogni paziente

```{r}
dat_ranked <- dat |> 
  group_by(patients) |> 
  mutate(region_rank = dense_rank(Region)) |> 
  ungroup()
```

Aggiungo variabili più leggibili

```{r}
# 1 pixel == 0.18 micron
dat_ranked$x_px <- dat_ranked$`X:X`
dat_ranked$y_px <- dat_ranked$`Y:Y`
dat_ranked$cell_type = dat_ranked$`neighborhood name`
```

Separo i dataset per le due regioni per ogni paziente, in modo da avere un dataset con la prima regione per tutti i pazienti e un altro per la seconda regione per tutti i pazienti.

```{r}
# 2. Extract the first dataset (Rank 1 regions)
dat_gr1 <- dat_ranked %>%
  filter(region_rank == 1) %>%
  select(-region_rank) # remove the helper column

# 3. Extract the second dataset (Rank 2 regions)
dat_gr2 <- dat_ranked %>%
  filter(region_rank == 2) %>%
  select(-region_rank) # remove the helper column
```

Salvo i due dataset separatamente
```{r}
#saveRDS(dat_gr1, file = "dataset/patients_region1.rds")
#saveRDS(dat_gr2, file = "dataset/patients_region2.rds")
rm(list=ls())
```

Rileggo i dati puliti
```{r}
dati <- readRDS("dataset/patients_region1.rds")
```

2. Split dataset into a list of dataframes (one per patient)

```{r}
patients_list <- split(dati, dati$patients)
```


3. Apply the function to each patient

```{r}
ppp_pats <- map(patients_list, function(pat) {
  # Note: Ensure 'cell_type' column exists in your data. 
  # If it is named 'ClusterID' or something else, update it here.
  celltypes <- as.factor(pat$cell_type)
  ppp_creation(pat, celltypes)
})

# Check the result
# ppp_pats is now a list of ppp objects, named by patient ID
print(names(ppp_pats))
```

There are NA in some marks and there are duplicated points. The pcf estimate isn't bothered with the duplicated points, but Lest and other functions may be.
The NA will be accounted for in the functions estimate.

# Omogeneus processes hypotesis

## Lest estimate
```{r}
Lest_list <- map(ppp_pats, function(pat){
  map(split(pat), function(ppp_mark){ 
    Lest(ppp_mark, correction = "best")})
})
```

### Plots

Data preparation
```{r}
plot_data <- imap_dfr(Lest_list, function(regions_list, patient_id) {
  imap_dfr(regions_list, function(fv_data, region_name) {
    
    # Extract data and calculate L(r) - r
    as.data.frame(fv_data) %>%
      select(r, iso) %>%
      mutate(
        PatientID = as.character(patient_id),
        Region = region_name,
        # The specific transformation you requested:
        L_centered = iso - r 
      )
  })
})
```

and actual plots
```{r}
# Create a list of plots
patient_plots <- plot_data %>%
  group_by(PatientID) %>%
  group_map(function(data, keys) {
    
    ggplot(data, aes(x = r, y = L_centered, color = Region)) +
      # Add reference line at 0 (Theoretical Random distribution)
      geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
      # Plot the curves
      geom_line(linewidth = 0.8, na.rm = TRUE) + 
      scale_color_brewer(palette = "Set1") +
      theme_bw() +
      labs(
        title = paste("Patient:", keys$PatientID),
        subtitle = "Centered L-function",
        y = "L(r) - r",
        x = "Radius (r)",
        color = "Region"
      )
  }, .keep = TRUE) # .keep=TRUE ensures grouping variable is retained if needed
```

```{r}
Lest_pplots <- Lest_plot(Lest_list)
walk(Lest_pplots, print)
```

## Pair correlation function

Verify if divisor_fun works as intended

```{r}
sapply(ppp_pats, divisor_h)
```

Hybrid estimator of the pcf. If the H0 of CSR in the clark evans test is rejected we use the new pcf estimator made by Baddeley ("a"), otherwise we use the standard corrected near zero.

```{r}
pcf_list <- map(ppp_pats, function(pat){
  map(split(pat), function(ppp_mark, divisor = divisor_h) {
    if(npoints(ppp_mark) < 5)
      return(NULL)
    pcf(ppp_mark, 
        correction = "best", 
        kernel = "rectangular",
        divisor = divisor, # it can be a chr or a fun that return the right chr value
        zerocor = "JonesFoster",
        bw = bw.diggle)
  })
})
```

### Plots

```{r}
pcf_plots <- pcf_plot(pcf_list)
walk(pcf_plots, print)
```


# FDA estimates

```{r}
Lfun_list <- map(ppp_pats, \(pat){
  Lest_smooth(pat, centrata = F)
})
```

