---
title: "merfish analysis"
author: "Valerio Reffo"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(anndataR)
library(ggspavis)
library(patchwork)
library(spatstat.geom)
library(spatstat)
library(sosta)
library(purrr)
library(readxl)
library(dplyr)
library(spatialFDA)
library(fda)
# survival analysis
library(survival)
library(survminer)

library(MerfishData)
```

Merfish dataset

```{r}
#spe <- MouseColonIbdCadinu2024()
#saveRDS(spe, file = "~/repositories/spatialPointProcess_clustering/dataset/Merfish_dataset/spatialExp_alldata.rds")

spe <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/Merfish_dataset/spatialExp_alldata.rds")
```

Rapporto micron per pixel: 1 pixel = 109 nanometri aka 109/1000 = 0.109 

```{r}
mpp <-  0.109
```

Per selezionare un campione specifico usare mouse_id e slice_id

```{r}
table(spe$sample_id)
table(spe$sample_type)
table(spe$mouse_id)
table(spe$mouse_id, spe$slice_id)
table(spe$slice_id, spe$technical_repeat_number)
```


```{r}
table(spe$sample_id, spe$slice_id, spe$mouse_id)
```


Identificativo di "campione" per le analisi successive
```{r}
spe$unique_id <- paste0(spe$mouse_id, "_", spe$sample_id, "_", spe$slice_id)
table(spe$unique_id) 
```

# Exploratory analysis

```{r}
meta_df <- as.data.frame(colData(spe))

# 2. Split the data frame by unique_id into a named list of data frames
df_list <- split(meta_df, meta_df$unique_id)

# 3. Use imap to iterate over the list (data) and its names (uid)
barp_list <- purrr::imap(df_list, \(data, uid) {
  ggplot(data, aes(x = tier1, fill = tier1)) +
    geom_bar() +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    ) +
    labs(
      title = "Cell Type (tier1) Distribution",
      subtitle = paste("Sample:", uid),
      x = "Tier 1 Cell Type",
      y = "Number of Cells"
    )
})
walk(barp_list, plot)
```


# Point process objects

Convert to point process class object as list of samples
```{r}
save_ppp_from_spe <- function(spe, win_list, 
                              out_dir = "~/repositories/spatialPointProcess_clustering/dataset/Merfish_dataset/samples_ppp", 
                              mpp = 0.109, 
                              mark_col = "tier1") {
  
  # 1. Ensure the output directory exists
  if (!dir.exists(out_dir)) {
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    message(paste("Created directory:", out_dir))
  }
  
  # Get all unique sample IDs from the SpatialExperiment
  sample_names <- unique(colData(spe)$unique_id)
  
  # 2. Iterate over each sample
  for (sample_name in sample_names) {
    message(paste("Processing:", sample_name))
    
    # 3. Retrieve the matching owin window
    win_final <- win_list[[sample_name]]
    
    # Safety check: does the window exist?
    if (is.null(win_final)) {
      warning(paste("No owin object found in windows_list for", sample_name, "- skipping."))
      next
    }
    
    # 4. Subset the SpatialExperiment for the current sample
    spe_sub <- spe[, colData(spe)$unique_id == sample_name]
    
    # Extract spatial coordinates
    coords <- spatialCoords(spe_sub)
    
    # Extract marks (e.g., cell types). Check if the column exists first.
    if (!mark_col %in% colnames(colData(spe_sub))) {
      warning(paste("Column", mark_col, "not found in colData. Creating unmarked ppp."))
      marks_factor <- NULL
    } else {
      marks_factor <- as.factor(colData(spe_sub)[[mark_col]])
    }
    
    # 5. Build the PPP object
    # spatstat will automatically drop points that fall outside the owin boundaries 
    # and throw a safe warning if it happens.
    my_ppp <- tryCatch({
      ppp(
        x = coords[, 1], 
        y = coords[, 2], 
        window = win_final,
        marks = marks_factor
      )
    }, error = function(e) {
      warning(paste("Error creating ppp for", sample_name, ":", e$message))
      return(NULL)
    })
    
    if (is.null(my_ppp)) next
    
    # 6. Rescale to microns
    if (is.na(mpp) || mpp == 1) {
      warning(paste("No valid MPP provided for", sample_name, "- leaving in pixels"))
      ppp_microns <- my_ppp
    } else {
      # In spatstat, 's' is the length of 1 new unit (micron) in old units (pixels).
      # If 1 pixel = 0.109 microns, then 1 micron = 1/0.109 pixels.
      ppp_microns <- rescale(my_ppp, s = 1/mpp, unitname = c("micron", "microns"))
    }
    
    # 7. Save the individual object to the specified directory
    file_path <- file.path(out_dir, paste0(sample_name, ".rds"))
    saveRDS(ppp_microns, file = file_path)
    
    message(paste("Saved successfully:", file_path))
  }
  
  message("All samples processed.")
}

# Example usage:
# save_ppp_from_spe(spe = spe,
#                   win_list = windows_list,
#                   mpp = 0.109,
#                   mark_col = "tier1") # change "type" to whatever your cell type column is named
```

```{r}
# input_dir <- "~/repositories/spatialPointProcess_clustering/dataset/Merfish_dataset/samples_ppp" # Change this to your actual folder path
# file_paths <- list.files(path = input_dir, pattern = "\\.rds$", full.names = TRUE)
# 
# ppp_list <- map(file_paths, readRDS)
# names(ppp_list) <- tools::file_path_sans_ext(basename(file_paths))

#saveRDS(ppp_list, file = "~/repositories/spatialPointProcess_clustering/dataset/Merfish_dataset/ppp_list_allsamples.rds")
ppp_list <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/Merfish_dataset/ppp_list_allsamples.rds")
```

Plot of the points inside and outside the masks

Plot to check if the mask is correct and the points are inside
```{r}
poinside_plot <- \(ppp, name){
  ggplot() +
    # 1. Plot the tissue mask (Extracts pixels on the fly)
    geom_raster(data = as.data.frame(Window(ppp)),
                aes(x = x, y = y), fill = "grey80") +

    # 2. Plot the valid cells (Extracts coordinates on the fly)
    geom_point(data = as.data.frame(ppp),
               aes(x = x, y = y, color = marks), size = 0.1, alpha = 0.5) +

    # Force standard Cartesian proportions so the image isn't squished
    coord_fixed() +
    theme_minimal() +
    labs(
      title = "Cell Alignment Check",
      subtitle = "Grey: Tissue Mask | Blue: Valid Cells | Red: Rejected Cells",
      x = "X Coordinate",
      y = "Y Coordinate"
    )
}
map(ppp_list, poinside_plot)
```

Solo fibroblasti e epiteliali
```{r}
poinside_plot <- \(ppp, name){
  
  # Extract the points and filter for only Fibroblast and Epithelial
  cell_data <- as.data.frame(ppp) |> 
    dplyr::filter(marks %in% c("Fibroblast", "Epithelial"))
  
  ggplot() +
    # 1. Plot the tissue mask
    geom_raster(data = as.data.frame(Window(ppp)),
                aes(x = x, y = y), fill = "grey80") +

    # 2. Plot ONLY the filtered valid cells
    geom_point(data = cell_data,
               aes(x = x, y = y, color = marks), size = 0.1, alpha = 0.7) +

    # 3. Define specific colors for your cell types
    scale_color_manual(values = c("Fibroblast" = "blue", "Epithelial" = "red")) +

    # Force standard Cartesian proportions so the image isn't squished
    coord_fixed() +
    theme_minimal() +
    labs(
      title = paste("Filtered Cells:", name), # Uses the list name
      subtitle = "Grey: Tissue Mask | Blue: Fibroblast | Red: Epithelial",
      x = "X Coordinate",
      y = "Y Coordinate",
      color = "Cell Type"
    )
}

# Use imap instead of map so 'name' gets passed correctly
filtered_plots <- imap(ppp_list, poinside_plot)
filtered_plots
```


# Second order statistic analysis

# Homogeneus ppp

## Lfunction estimate

Between Fibroblast and Epithelial
```{r}
Lcross_FibEp <- map(ppp_list, \(ppp_obj) {
     # Calculate Lcross
     L <- Lcross(ppp_obj, i = "Fibroblast", j = "Epithelial", correction = "border", rmax = 12)
     L
})
```

Check for NA
```{r}
na_summary_per_sample <- imap_dfr(Lcross_FibEp, \(fv_obj, sample_name) {
  
  # Handle cases where the Lcross calculation returned NULL
  if (is.null(fv_obj)) {
    return(tibble(
      sample = sample_name,
      r_na = NA_integer_,
      theo_na = NA_integer_,
      border_na = NA_integer_,
      status = "Skipped (NULL)"
    ))
  }
  
  # Convert the fv object to a standard dataframe
  df <- as.data.frame(fv_obj)
  
  # Count NAs for each specific column
  tibble(
    sample = sample_name,
    r_na = sum(is.na(df$r)),
    theo_na = sum(is.na(df$theo)),
    border_na = sum(is.na(df$border)),
    status = "Computed"
  )
})

# View the per-sample breakdown
print(na_summary_per_sample)

# 2. Count NAs for EVERYTHING (Grand Totals)
grand_totals <- na_summary_per_sample |> 
  summarise(
    total_r_NAs = sum(r_na, na.rm = TRUE),
    total_theo_NAs = sum(theo_na, na.rm = TRUE),
    total_border_NAs = sum(border_na, na.rm = TRUE),
    total_failed_samples = sum(status == "Skipped (NULL)")
  )

# View the total counts
print(grand_totals)
```

## Plot of the Lcross curves

```{r}
plot_fv_list <- function(list_of_funcs, spe, correction = "border") {
  
  # 1. Extract unique mapping of unique_id to sample_type from the SPE object
  meta_map <- as.data.frame(colData(spe)) |> 
    dplyr::select(unique_id, sample_type) |> 
    dplyr::distinct()
  
  # 2. Run the map_dfr block dynamically on whatever list you provide
  df_lcross <- map_dfr(names(list_of_funcs), function(slide_name) {
    
    # Extract the specific L-function object from the provided list
    L_obj <- list_of_funcs[[slide_name]]
    
    as.data.frame(L_obj) |> 
      mutate(
        Slide = slide_name,
        # Use .data[[correction]] to dynamically grab the 'border' column (or 'iso', etc.)
        L_r_minus_r = .data[[correction]] - r,  
        
        # Extract the marks names exactly as you designed
        mark_i = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][1]),
        mark_j = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][2])
      )
  })
  
  # 3. Join the sample_type metadata to our plot dataframe
  df_lcross <- df_lcross |> 
    left_join(meta_map, by = c("Slide" = "unique_id"))
  
  # 4. Plot with color mapped to sample_type
  p <- ggplot(df_lcross, aes(x = r, y = L_r_minus_r, group = Slide, color = sample_type)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(alpha = 0.7, linewidth = 0.6) + 
    theme_minimal() +
    labs(
      title = paste0("L-cross Function: ", df_lcross$mark_i[1], " vs ", df_lcross$mark_j[1]),
      subtitle = "Centered L-function (L(r) - r). Positive = Attraction, Negative = Repulsion",
      x = "Radius (r)",
      y = "L(r) - r",
      color = "Sample Type"
    ) +
    theme(legend.position = "right")
  
  return(p)
}
```


```{r}
(plotEp12 <- plot_fv_list(list_of_funcs = Lcross_FibEp, spe = spe, correction = "border"))
```

```{r}
fsmooth <- function(fcross_list, fun, M = 6, genLfun.fd = 4, centrata = F, ...)
{
  # 1. Calculate raw L-functions
  r <- fcross_list[[1]]$r
  mat_fun <- sapply(fcross_list, function(x) x$border)
  
  # Ensure column names exist
  if(is.null(colnames(mat_fun))) colnames(mat_fun) <- names(fcross_list)

  # --- FIX: HANDLE NAs BEFORE SMOOTHING ---
  for(i in 1:ncol(mat_fun)) {
    # If the ENTIRE curve is NaN (because a cell type was missing)
    if(all(is.na(mat_fun[, i]) | is.nan(mat_fun[, i]))) {
      mat_fun[, i] <- r # We set it to 'r' so it centers to 0 later
    } 
    # If only PART of the curve is NA, we interpolate the missing points
    else if(any(is.na(mat_fun[, i]) | is.nan(mat_fun[, i]))) {
      # Approximates the missing points based on surrounding points
      mat_fun[, i] <- zoo::na.approx(mat_fun[, i], na.rm = FALSE, rule = 2)
    }
  }
  # ----------------------------------------

  # 2. Apply centering
  if(centrata == TRUE){
    mat_fun <- sweep(mat_fun, MARGIN = 1, STATS = r, FUN = "-")
  }

  # 3. Create Basis and Smooth
  K <- length(r) + M - 2
  genbasis <- create.bspline.basis(range(r), K, M, r)
  genfdPar <- fdPar(genbasis, genLfun.fd, lambda = 1e-11)
  
  # NOW it will just smooth perfectly
  genfdSmooth <- smooth.basis(r, mat_fun, genfdPar)
  Genfun.fd <- genfdSmooth$fd
  
  # 4. Set Names
  fdn_dist <- paste("Distanza (da 0 a", max(r), "micron)")
  fdnames <- list(fdn_dist,
                  "Sample" = colnames(mat_fun),
                  "Funzione")
  Genfun.fd$fdnames <- fdnames
  
  return(list(fd = Genfun.fd, fdPar = genfdPar, 
              basis = genbasis, mat = mat_fun, r = r))
}

# Run your smoothing
Lfun_FibEp <- fsmooth(Lcross_FibEp, centrata = T)

# Now it will plot without crashing
plot(Lfun_FibEp$fd)
```


```{r}
plot(Lfun_FibEp$fd)
lines(c(0,100), c(0,0), col = "gold", lwd = 2)
```

### Functional PCA

Calcoliamo fpca per 4 componenti e mostriamo il risultato sulla rotazione VARIMAX

```{r}
fpca <- pca.fd(Lfun_FibEp$fd, nharm = 4, Lfun_FibEp$fdPar)
rownames(fpca$scores) <- names(Lcross_FibEp) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca)
```

```{r}
plot.pca.fd(fpcaVarmx, cex.main=0.9)
```

## Clustering

Applichiamo due metodi di clustering sugli scores delle prime 4 componenti.
Clustering gerarchico con metodo Ward e kmeans

```{r}
set.seed(123)
fpca_km <- kmeans(fpca$scores[,c(1,4)], 4)
fpca_hc <- hclust(dist(fpca$scores[,c(1,4)]), method = "ward.D2")
memb <- cutree(fpca_hc, k = 4)
```

Grafico del dendogramma per il clustering gerarchico, scegliamo di dividere in 5 cluster.

```{r}
dhc <- as.dendrogram(fpca_hc)
table(kmean = fpca_km$cluster)
table(ward.d2 = memb)
```

```{r}
plot(fpca$scores[,1], fpca$scores[,2], col = fpca_km$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca$scores[,1], fpca$scores[,2], col = memb, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

## Check if there is correspondence between sample_type (weeks of patology) and clusters
```{r}
colData(spe)$fpca_cluster <- as.factor(fpca_km$cluster[colData(spe)$unique_id])

# Check to make sure it worked
table(colData(spe)$fpca_cluster, useNA = "ifany", spe$sample_type)
```

```{r}
sample_summary <- as.data.frame(colData(spe)) |>
  dplyr::select(unique_id, sample_type, fpca_cluster) |>
  dplyr::distinct()

# 2. Create a Cross-Tabulation (Contingency Table)
# This will show you exactly how many samples of each type fall into each cluster
cat_table <- table(sample_summary$sample_type, sample_summary$fpca_cluster)
print("Count of samples per Sample Type and Cluster:")
print(cat_table)
```




# Between Fibroblast and Immune

```{r}
Lcross_FibImm <- map(ppp_list, \(ppp_obj) {
     # Calculate Lcross
     L <- Lcross(ppp_obj, i = "Fibroblast", j = "Immune", correction = "border", rmax = 12)
     L
})
```

Check for NA
```{r}
na_summary_per_sample <- imap_dfr(Lcross_FibEp, \(fv_obj, sample_name) {
  
  # Handle cases where the Lcross calculation returned NULL
  if (is.null(fv_obj)) {
    return(tibble(
      sample = sample_name,
      r_na = NA_integer_,
      theo_na = NA_integer_,
      border_na = NA_integer_,
      status = "Skipped (NULL)"
    ))
  }
  
  # Convert the fv object to a standard dataframe
  df <- as.data.frame(fv_obj)
  
  # Count NAs for each specific column
  tibble(
    sample = sample_name,
    r_na = sum(is.na(df$r)),
    theo_na = sum(is.na(df$theo)),
    border_na = sum(is.na(df$border)),
    status = "Computed"
  )
})

# View the per-sample breakdown
print(na_summary_per_sample)

# 2. Count NAs for EVERYTHING (Grand Totals)
grand_totals <- na_summary_per_sample |> 
  summarise(
    total_r_NAs = sum(r_na, na.rm = TRUE),
    total_theo_NAs = sum(theo_na, na.rm = TRUE),
    total_border_NAs = sum(border_na, na.rm = TRUE),
    total_failed_samples = sum(status == "Skipped (NULL)")
  )

# View the total counts
print(grand_totals)
```

## Plot of the Lcross curves

```{r}
(plot2 <- plot_fv_list(list_of_funcs = Lcross_FibImm, spe = spe, correction = "border"))
```

```{r}
fsmooth <- function(fcross_list, fun, M = 6, genLfun.fd = 4, centrata = F, ...)
{
  # 1. Calculate raw L-functions
  r <- fcross_list[[1]]$r
  mat_fun <- sapply(fcross_list, function(x) x$border)
  
  # Ensure column names exist
  if(is.null(colnames(mat_fun))) colnames(mat_fun) <- names(fcross_list)

  # --- FIX: HANDLE NAs BEFORE SMOOTHING ---
  for(i in 1:ncol(mat_fun)) {
    # If the ENTIRE curve is NaN (because a cell type was missing)
    if(all(is.na(mat_fun[, i]) | is.nan(mat_fun[, i]))) {
      mat_fun[, i] <- r # We set it to 'r' so it centers to 0 later
    } 
    # If only PART of the curve is NA, we interpolate the missing points
    else if(any(is.na(mat_fun[, i]) | is.nan(mat_fun[, i]))) {
      # Approximates the missing points based on surrounding points
      mat_fun[, i] <- zoo::na.approx(mat_fun[, i], na.rm = FALSE, rule = 2)
    }
  }
  # ----------------------------------------

  # 2. Apply centering
  if(centrata == TRUE){
    mat_fun <- sweep(mat_fun, MARGIN = 1, STATS = r, FUN = "-")
  }

  # 3. Create Basis and Smooth
  K <- length(r) + M - 2
  genbasis <- create.bspline.basis(range(r), K, M, r)
  genfdPar <- fdPar(genbasis, genLfun.fd, lambda = 1e-11)
  
  # NOW it will just smooth perfectly
  genfdSmooth <- smooth.basis(r, mat_fun, genfdPar)
  Genfun.fd <- genfdSmooth$fd
  
  # 4. Set Names
  fdn_dist <- paste("Distanza (da 0 a", max(r), "micron)")
  fdnames <- list(fdn_dist,
                  "Sample" = colnames(mat_fun),
                  "Funzione")
  Genfun.fd$fdnames <- fdnames
  
  return(list(fd = Genfun.fd, fdPar = genfdPar, 
              basis = genbasis, mat = mat_fun, r = r))
}

# Run your smoothing
Lfun_FibEp <- fsmooth(Lcross_FibEp, centrata = T)

# Now it will plot without crashing
plot(Lfun_FibEp$fd)
```

```{r}
Lfun_FibImm <- fsmooth(Lcross_FibImm, centrata = T)
```

```{r}
plot(Lfun_FibImm$fd)
lines(c(0,100), c(0,0), col = "gold", lwd = 2)
```

### Functional PCA

Calcoliamo fpca per 4 componenti e mostriamo il risultato sulla rotazione VARIMAX

```{r}
fpca_2 <- pca.fd(Lfun_FibImm$fd, nharm = 4, Lfun_FibImm$fdPar)
rownames(fpca_2$scores) <- names(Lcross_FibImm) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca_2)
```

```{r}
plot.pca.fd(fpcaVarmx, cex.main=0.9)
```

## Clustering

Applichiamo due metodi di clustering sugli scores delle prime 4 componenti.
Clustering gerarchico con metodo Ward e kmeans

```{r}
set.seed(123)
fpca_km <- kmeans(fpca_2$scores[,c(1,4)], 4)
fpca_hc <- hclust(dist(fpca_2$scores[,c(1,4)]), method = "ward.D2")
memb <- cutree(fpca_hc, k = 4)
```

Grafico del dendogramma per il clustering gerarchico, scegliamo di dividere in 5 cluster.

```{r}
dhc <- as.dendrogram(fpca_hc)
table(kmean = fpca_km$cluster)
table(ward.d2 = memb)
```

```{r}
plot(fpca_2$scores[,1], fpca_2$scores[,2], col = fpca_km$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca_2$scores[,1], fpca_2$scores[,2], col = memb, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

## Check if there is correspondence between sample_type (weeks of patology) and clusters
```{r}
colData(spe)$fpca_cluster2 <- as.factor(fpca_km$cluster[colData(spe)$unique_id])

# Check to make sure it worked
table(colData(spe)$fpca_cluster2, useNA = "ifany", spe$sample_type)
```

```{r}
sample_summary <- as.data.frame(colData(spe)) |>
  dplyr::select(unique_id, sample_type, fpca_cluster2) |>
  dplyr::distinct()

# 2. Create a Cross-Tabulation (Contingency Table)
# This will show you exactly how many samples of each type fall into each cluster
cat_table <- table(sample_summary$sample_type, sample_summary$fpca_cluster2)
print("Count of samples per Sample Type and Cluster:")
print(cat_table)
```