---
title: "EDA - all 4 samples per patient"
author: "Valerio Reffo"
date: "2026-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(spatstat)
library(tidyverse)
library(purrr)
library(fda)
library(readxl)
library(pals)
library(knitr)
library(tidyr)
```


```{r}
dati = read_csv("dataset/CRC_clusters_neighborhoods_markers.csv")
```

Reduce the number of clusters
```{r}
unique(dati$ClusterName)
```

"tumor cells"  

"CD4+ T cells CD45RO+"       
"CD4+ T cells GATA3+"        
"CD3+ T cells" 
"CD8+ T cells"               
"CD4+ T cells"              
"Tregs"  

"B cells"  
"plasma cells" 

"CD68+ macrophages"         
"CD68+ macrophages GzmB+"    
"CD68+CD163+ macrophages"    
"CD11b+CD68+ macrophages"    
"CD163+ macrophages"      
"CD11c+ DCs"                 
"CD11b+ monocytes"   
"granulocytes" 

"adipocytes"   
"nerves"                     
"smooth muscle"             
"stroma"
"vasculature"
"lymphatics"  

"dirt"    

"NK cells"                   

"immune cells"   

"tumor cells / immune cells" 

"immune cells / vasculature" 

"undefined"                  


```{r}
dati <- dati |> 
  mutate(broad_type = case_when(
    ClusterName %in% c("CD4+ T cells CD45RO+", "CD8+ T cells", "Tregs", 
                     "CD4+ T cells", "CD4+ T cells GATA3+", "CD3+ T cells") ~ "T cells",
    
    ClusterName %in% c("CD68+CD163+ macrophages", "CD11b+CD68+ macrophages", 
                     "CD68+ macrophages GzmB+", "CD68+ macrophages", 
                     "CD163+ macrophages", "CD11b+ monocytes", 
                     "CD11c+ DCs", "granulocytes")                          ~ "Myeloid",
    
    ClusterName == "B cells"                                                ~ "B cells",
    ClusterName == "plasma cells"                                           ~ "plasma cells",
    ClusterName == "NK cells"                                               ~ "NK cells",
    ClusterName == "tumor cells"                                            ~ "Tumor",
    ClusterName %in% c("stroma", "adipocytes", "smooth muscle", "nerves")   ~ "Stroma",
    ClusterName %in% c("vasculature", "lymphatics")                         ~ "Vasculature",
    ClusterName == "immune cells"                                           ~ "Immune cells",
    ClusterName == "tumor cells / immune cells"                             ~ "Tumor/immune",
    ClusterName == "immune cells / vasculature"                             ~ "Immune/vascular",
    ClusterName == "undefined"                                              ~ "Undefined",
    ClusterName == "dirt"                                                   ~ "Dirt"
  ))

table(dati$broad_type)
```

We have 4 samples for each patient

```{r}
samples_list <- split(dati, dati$`File Name`)
length(samples_list)
```

Create a single ppp for each sample

```{r}
marks_name <- c("ClusterName", "neighborhood name", "broad_type")
ppp_pats <- map(samples_list, function(pat) {
  # Note: Ensure 'cell_type' column exists in your data. 
  # If it is named 'ClusterID' or something else, update it here.
  ppp_creation(pat, marks_name, "X:X", "Y:Y")
})

# Check the result
# ppp_pats is now a list of ppp objects, named by patient ID
print(names(ppp_pats))
```

Patients data

```{r}
pat_data <- read_excel("dataset/CRC_TMAs_patient_annotations.xlsx", n_max = 35)
pat_data <- pat_data |>  mutate(Patient = as.character(Patient))
head(pat_data)
```

Let's create an object that can accomodate every kind of object we would need to analyze the dataset,
for each sample we can add info about, the object is an Hyper Data frame

```{r}
ppp_hf <- as.hyperframe(ppp = as.solist(ppp_pats))
ppp_hf <- cbind(ppp_hf, sample_name = unique(dati$`File Name`))
```

Let's add all the info about the patients
```{r}
# 1. Extract unique metadata for each sample (File Name)
# We select the key and the columns you want to add
metadata <- dati |> 
  dplyr::select(`File Name`, 
    Region, 
    groups, 
    patients, 
    spots) |> 
  # Keep only one row per unique 'File Name'
  distinct(`File Name`, .keep_all = TRUE)

# 2. Align the metadata with the hyperframe order
# We match 'sample_name' from the hyperframe to 'File Name' in the metadata
matching_indices <- match(ppp_hf$sample_name, metadata$`File Name`)

# 3. Add the columns to the Hyperframe
# We assign all columns except the key ("File Name")
cols_to_add <- setdiff(colnames(metadata), "File Name")
ppp_hf[, cols_to_add] <- metadata[matching_indices, cols_to_add]

# Verify the result
print(ppp_hf)
```


```{r}
ppp_df <- dati |> 
  dplyr::select(`File Name`, 
    Region, 
    groups, 
    patients, 
    spots,
    ClusterName,
    `neighborhood name`,
    broad_type,
    `X:X`,
    `Y:Y`
    ) |> 
  mutate(
    sample = `File Name`,
    patients = as.character(patients),
    x = `X:X`,
    y = `Y:Y`
  )
ppp_df <- left_join(ppp_df, pat_data, by = c("patients" = "Patient"))
```


Save the object to work with
```{r}
saveRDS(ppp_df, file = "dataset/data_for_spatialplot.rds")
saveRDS(ppp_hf, file = "dataset/hyperframe_140samples_integrated_info.rds")
```


# Plot

by patients, of course it's wrong
```{r}

ggplot(ppp_df, aes(x = x, y = y, color = ClusterName)) +
  # Use a small size and some transparency (alpha) because you have 250k points
  geom_point(size = 0.1, alpha = 0.5) +
  
  # This creates a separate plot for each patient
  facet_wrap(~ patients, scales = "free") +
  
  # Essential for spatial data: keeps the 1:1 aspect ratio of the tissue
  #coord_fixed() +
  
  # Clean up the look
  theme_minimal() +
  labs(title = "Spatial Distribution of Clusters by Patient",
       x = "X Coordinate",
       y = "Y Coordinate",
       color = "Cell Type") +
  
  # Optional: customize the legend if there are too many categories
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

```

by sample
```{r}
group_pats <- ppp_df |>
  distinct(patients) |>
  mutate(page = ntile(patients, 4)) # Divides patients into 4 equal-sized groups
```


```{r}
all_plots <- ppp_df |> 
  left_join(group_pats, by = "patients") |> 
  group_split(page) |>
  map(\(page) {
    
    num_clusters <- length(unique(ppp_df$`neighborhood name`))
    my_colors <- as.vector(pals::glasbey(num_clusters))
    # .x represents each chunk of data
    ggplot(page, aes(x = x, y = y, color = `neighborhood name`)) +
      geom_point(size = 0.1, alpha = 0.6) +
      facet_wrap(~ sample, scales = "free", ncol = 6) +
      scale_color_manual(values = my_colors) + 
      #coord_fixed() +
      theme_minimal() +
      theme(
        legend.position = "left",             # Sposta a sinistra
        legend.text = element_text(size = 6),  # Testo piccolo per risparmiare spazio
        legend.title = element_text(size = 8), # Titolo piccolo
        legend.key.size = unit(0.2, "cm"),     # Rimpicciolisce i quadratini della legenda
        legend.spacing.x = unit(0.1, "cm")     # Riduce lo spazio tra simbolo e testo
      ) +
      guides(color = guide_legend(
        override.aes = list(size = 3, alpha = 1), # Simboli grandi nella legenda
        ncol = 1                                  # Una sola colonna per occupare meno larghezza
      )) +
      
      labs(title = paste("Patient Spatial Distribution"),
           x = "X Coordinate", y = "Y Coordinate")
  })

iwalk(all_plots, ~{
  ggsave(
    filename = paste0("Patient_Page_CN", .y, ".png"), 
    plot = .x, 
    width = 20,     # Dimensioni grandi per mantenere la risoluzione
    height = 15, 
    dpi = 300       # Alta qualità
  )
})# Now 'all_plots' is a list of 4 ggplot objects.

# Or save them all to a PDF in one line
walk(all_plots, print) # 'walk' is like map but for actions (printing)
```


# Quality check and filtering of samples with too low counts

```{r}

all_plots <- ppp_df |> 
  left_join(group_pats, by = "patients") |> 
  group_split(page) |>
  map(\(page) {
    
    num_clusters <- length(unique(ppp_df$broad_type))
    my_colors <- as.vector(pals::glasbey(num_clusters))
    # .x represents each chunk of data
    ggplot(page, aes(x = broad_type, fill = broad_type)) +
      geom_bar() +
      facet_wrap(~ sample, scales = "free_y") + # Using 'sample' as requested
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "none" # Hide legend since X axis has the labels
      ) +
      labs(
        x = "Broad Cell Type", 
        y = "Cell Count", 
        title = "Distribution of Cell Lineages per Sample"
      )
  })
walk(all_plots, print)
```


```{r}
library(gt) # nice tables i guess

summary_df <- ppp_df |> 
    group_by(patients) |> 
    count(sample, broad_type)

# 2. Split by patient and map to gt tables
table_list <- summary_df |>
  group_split(patients) |>
  map(function(df) {
    df |>
      ungroup() |>
      #select(-patients) |>  # Remove patient column (it will go in the title)
      gt() |>
      tab_header(
        title = paste("Patient ID:", unique(df$patients))
      )
  })

# View the table for the first patient
walk(table_list, print)
```

```{r}
target_types <- c("Tumor", "Myeloid", "T cells")

# 1. Generate the counts dataframe (as established previously)
counts_df <- map_dfr(ppp_hf$ppp, function(x) {
  current_types <- x$marks$broad_type 
  target_types %>%
    set_names() %>%
    map_int(~ sum(current_types == .x, na.rm = TRUE))
})
# Add sample names for identification
counts_df$sample_name <- ppp_hf$sample_name

# 2. Extract the top N samples for EACH target type
# This creates a list where each element corresponds to a cell type
top_performers <- map(target_types, function(type) {
  
  counts_df %>%
    # Sort by the current type in descending order
    arrange(desc(.data[[type]])) %>%
    # Take the top 5 (change 'n' to whatever number you need)
    slice_head(n = 5) %>%
    # Add a column to indicate which type was used for sorting
    mutate(Sorted_By = type)
})

# Name the list elements for easy access
names(top_performers) <- target_types

# --- View the results ---

# See the samples with the most Tumor cells (and their other counts)
print("Top samples for Tumor:")
print(top_performers[["Tumor"]])

# See the samples with the most Myeloid cells
print("Top samples for Myeloid:")
print(top_performers[["Myeloid"]])

# See the samples with the most T cells
print("Top samples for T cells:")
print(top_performers[["T cells"]])



```


