---
title: "All samples analysis"
author: "Valerio Reffo"
date: "2026-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(anndataR)
library(ggspavis)
library(patchwork)
library(spatstat.geom)
library(spatstat)
library(sosta)
library(readxl)
library(purrr)
library(dplyr)
library(spatialFDA)
library(fda)
# survival analysis
library(survival)
library(survminer)
```

Path of the HoverNet data

```{r}
data_path <- "/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/h5ad"
img_path <- "/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/thumb"
```

# Micron per Pixel extraction

Extract from xlsx file the info about mpp conversion
```{r}
xlsx_path <- "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/TCGA_dataset/TCGA_data/image_metadata_TCGA.xlsx"

# 2. Read the excel file
metadata <- read_excel(xlsx_path)

# 2b. Correct row 2 because that slide is 40x, not 20x
metadata <- metadata |>
  mutate(
    `Image Name` = sub("\\.[a-zA-Z0-9]+$", "", `Image Name`),
    # Use grepl to match the start of the image name
    `Objective Power` = if_else(grepl("TCGA-13-A5FU", `Image Name`), "40", `Objective Power`),
    AppMag            = if_else(grepl("TCGA-13-A5FU", `Image Name`), "40", AppMag),
    MPP               = if_else(grepl("TCGA-13-A5FU", `Image Name`), "0.252", MPP),
    MPP_X             = if_else(grepl("TCGA-13-A5FU", `Image Name`), "0.252", MPP_X),
    MPP_Y             = if_else(grepl("TCGA-13-A5FU", `Image Name`), "0.252", MPP_Y)
  )

# # 3. Filter for your 6 samples and extract the Image Name and MPP columns
# # (Note: using backticks because the column name has a space)
# mpp_values <- metadata |> 
#   dplyr::mutate(core_id = sub("\\.svs$", "", `Image Name`),
#     patientID = substr(`Image Name`, 1, 12)) |> 
#   dplyr::select(`Image Name`, patientID, MPP, core_id) |> 
#   dplyr::mutate(MPP = as.numeric(MPP))
# 
# # 4. Convert it into a named vector so you can easily use it later
# # (e.g., mpp_vector["TCGA-42-2587...svs"] will return its specific MPP)
# mpp_vector <- setNames(mpp_values$MPP, mpp_values$`Image Name`)

# Print the result
# print(mpp_vector)
```

## Windows

HoverNet masks saving and reading

```{r}
# win_hovernet <- imap(spe_list, \(spe, name) {
# 
#   base_dir <- "/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/mask"
#   png_name <- paste0(sub("\\.[a-zA-Z0-9]+$", "", name), ".png")
#   img_file <- file.path(base_dir, png_name)
#   
#   # Add a safety check so you get a clear error if the file is truly missing
#   if (!file.exists(img_file)) {
#     stop(paste("Cannot find image file:", img_file))
#   }
#   
#   img <- imager::load.image(img_file)
#   
#   return(img)
# })
# 
# # saveRDS(win_hovernet, "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/TCGA_data/window_masks_HoverNet_fullsamples.rds")
# 
# 
# win_hovernet <- readRDS("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/TCGA_data/window_masks_HoverNet_fullsamples.rds")
```

## Metadata 

```{r}
metadata <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/TCGA_dataset/TCGA_data/survival_dataset_&_metadata.rds")
```


Function to transform from anndata to spatialExperiment
```{r}
ann2spe <- \(data_path, 
             spat_coords_name = c("x", "y"), 
             img_path = NULL, 
             mpp_value = NULL, ...) {
  ann <- anndataR::read_h5ad(data_path)
  #object to build SpatialExperiment
  counts <- t(ann$X) # transpose it
  cd <- ann$obs
  meta <- ann$uns
  spat <- ann$obsm$spatial
  colnames(spat) <- spat_coords_name
  cd <- cbind(cd, spat)
  spe <- SpatialExperiment::SpatialExperiment(assays = base::list("counts" = counts), 
                                              colData = cd,
                                              metadata = meta,
                                              spatialCoordsNames = spat_coords_name,
                                              imageSources = img_path, ...)
  metadata(spe)$MPP <- mpp_value
  spe
}
```

Reading all the samples togheter

```{r}

# spe_list <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/spe_list_107samples_TCGA-OV.rds")

# 2. Get a list of ALL files in that folder ending with .h5ad
# full.names = TRUE ensures we get the complete file path, not just the file name
# h5_files <- list.files(path = data_path, pattern = "\\.h5ad$", full.names = TRUE)
# 
# # Optional: Set the names of the vector to the file names (without the path) 
# # so your final list of SPE objects is neatly named!
# names(h5_files) <- sub("\\.h5ad$", "", basename(h5_files))
# 
# # 3. Loop through all the files and apply your function
# spe_list <- imap(h5_files, function(current_h5_file, file_name) {
#   
#   # Optional: If your images are in another folder, you can construct the img_path here
#   current_img_file <- file.path(img_path, paste0(file_name, ".png")) 
# 
#   # B. Look up the exact MPP value for this specific core_id
#   current_mpp <- unname(mpp_values$MPP[mpp_values$core_id == file_name])
#   
#   # Optional check: If MPP is missing from the tibble, warn the user
#   if (is.na(current_mpp)) {
#     message("Warning: No MPP value found for ", file_name, ". Setting to NA.")
#   }
#   
#   # 4. Use tryCatch to prevent the loop from crashing if one file is broken
#   tryCatch({
#     
#     # Call your custom function!
#     # (If you don't have images/mpp values for this batch, leave them as NULL)
#     spe <- ann2spe(
#       data_path = current_h5_file,
#       img_path = current_img_file,      # Update this if you have images
#       mpp_value = current_mpp      # Update this if you have mpp values
#     )
#     
#     return(spe)
#     
#   }, error = function(e) {
#     # If a file fails to load, print an error message and return NULL
#     message("Failed to load: ", basename(current_h5_file))
#     message("Error: ", e$message)
#     return(NULL)
#   })
#   
# }) |> 
#   # 5. Drop any NULLs (files that failed to load)
#   compact()

#saveRDS(spe_list, "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/spe_list_107samples_TCGA-OV.rds")
```

# Checking for bad samples

Sample spe_list 
```{r}
# set.seed(123)
# thumbs_samples <- imap(spe_list, \(spe, name) {
#   subspe <- spe[, sample(ncol(spe), 10000)]
#   plot <- ggplot(spatialCoords(subspe), aes(x = x, y = y)) +
#     # 2. Plot the valid cells (Extracts coordinates on the fly)
#     geom_point(color = "blue", size = 0.1, alpha = 0.5) +
#     # Force standard Cartesian proportions so the image isn't squished
#     coord_fixed() +
#     theme_minimal() +
#     labs(
#       title = name)
#   plot
# })
# thumbs_samples
```

Images with apparently more copy of the same tissue
```{r}
double_ids <- c(
  "TCGA-13-A5FT",
  "TCGA-25-1631",
  "TCGA-25-1632",
  "TCGA-25-1633",
  "TCGA-25-1877",
  "TCGA-42-2582",
  "TCGA-42-2588",
  "TCGA-42-2589",
  "TCGA-42-2590",
  "TCGA-42-2591",
  "TCGA-42-2593",
  "TCGA-59-A5PD",
  "TCGA-OY-A56P",
  "TCGA-OY-A56Q",
  "TCGA-VG-A8LO",
  "TCGA-VG-A8LO"   # doppio perché appare 2 volte 01-DX1 e 02-DX2
)

#saveRDS(double_ids, file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/bad_samples_fromthe107.rds")
```

Remove bad samples from the spe_list
Add the sample_id with the short name of the sample to identify the patient and then filter the samples in which I have similar slice of tissue
and I can't understand if they are replicate or adiacent samples, and in which direction is the adiacence.

```{r}
# spe_list <- imap(spe_list, \(spe, name) {
#   spe$sample_id <- NULL
#   spe$sample_id <- substr(name, 1, 12)
#   spe
# })
# spe_list <- spe_list[!substr(names(spe_list), 1, 12) %in% double_ids]
```

16 samples removed

Dimension of spe_list 
```{r}
# format(object.size(spe_list), units = "MB")
```

Resave the thing and divide it in 2 half
```{r}
#saveRDS(spe_list, file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/spe_list_91_good_samples.rds")

# spe_list <- readRDS("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/spe_list_91_good_samples.rds")
```

```{r}
# length(spe_list)
# (1:91)[1:floor(length(spe_list)/2)]
# (1:91)[-(1:floor(length(spe_list)/2))]

# # first 45 samples
# spe_list[1:floor(length(spe_list)/2)] |> 
#   saveRDS("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/spe_list_first_45_good_samples.rds")
# 
# # second 46 samples
# spe_list[-(1:floor(length(spe_list)/2))] |> 
#   saveRDS("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/spe_list_second_46_good_samples.rds")

```


# ppp objects creation 

```{r}
save_ppp_individually <- function(spe_list, mask_hovernet, metadata, 
                                  out_dir = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/samples_ppp") {
  
  # 1. Ensure the output directory exists
  if (!dir.exists(out_dir)) {
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    message(paste("Created directory:", out_dir))
  }
  
  # 2. Iterate over each sample
  for (sample_name in names(spe_list)) {
    message(paste("Processing:", sample_name))
    
    spe  <- spe_list[[sample_name]]
    mask <- mask_hovernet[[sample_name]]
    
    # Safety check
    if (is.null(mask)) {
      warning(paste("No mask found for", sample_name, "- skipping."))
      next
    }
    
    # 3. Extract metadata for the current sample
    meta_row <- metadata |> 
      dplyr::filter(`Image Name` == sample_name) |> 
      dplyr::select(AppMag, MPP)
    
    # Safely default to 40 and 1 if metadata is missing or NA
    if (nrow(meta_row) == 0) {
      app_mag <- 40
      mpp     <- 1
    } else {
      app_mag <- tidyr::replace_na(as.numeric(meta_row$AppMag[1]), 40)
      mpp     <- tidyr::replace_na(as.numeric(meta_row$MPP[1]), 1)
    }
    
    # A. Extract logical matrix and transpose to [y, x]
    mask_mat <- mask[,,1,1] > 0.5  
    mask_spatstat <- t(mask_mat)                 
    
    # B. Create the thumbnail window
    # Note: If width()/height() are custom functions, ensure they are loaded. 
    # Otherwise, you can use dim(mask)[1] and dim(mask)[2]
    win_thumb <- owin(
      xrange = c(0, dim(mask)[1]), 
      yrange = c(0, dim(mask)[2]), 
      mask = mask_spatstat
    )
    
    # C. Calculate exact scale
    exact_scale <- app_mag / 1.25
    
    # D. Stretch the window
    win_final <- affine(win_thumb, mat = diag(c(exact_scale, exact_scale)))
    
    # E. Build the PPP object
    coords <- spatialCoords(spe)
    my_ppp <- ppp(
      x = coords[, "x"], 
      y = coords[, "y"], 
      window = win_final,
      marks = as.factor(spe$type)
    )
    
    # F. Rescale
    if (mpp == 1) {
      warning(paste("No MPP found for", sample_name, "- leaving in pixels"))
    }
    
    ppp_microns <- rescale(my_ppp, s = 1/mpp, unitname = c("micron", "microns"))
    
    # 4. Save the individual object to the specified directory
    file_path <- file.path(out_dir, paste0(sample_name, ".rds"))
    saveRDS(ppp_microns, file = file_path)
    
    message(paste("Saved successfully:", file_path))
  }
  
  message("All samples processed.")
}

# # Run the function:
# save_ppp_individually(spe_list, win_hovernet, metadata)
```

Iterative reading and acting with ppp samples
```{r}
# sample_names <- names(spe_list)
# saveRDS(sample_names, file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/samples_name.rds")
```


```{r}
# path_dir = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/samples_ppp"
# map(sample_names, \(sample_name){
#     file_path <- file.path(path_dir, paste0(sample_name, ".rds"))
#     ppp_sample <- readRDS(file_path)
#     
#     # 1. Save the original marks vector
#     original_marks <- marks(ppp_sample)
#     
#     # 2. Create the new merged vector
#     target_levels <- c("benign epithelial", "stromal")
#     
#     if (any(target_levels %in% levels(original_marks))) {
#       merged_marks <- forcats::fct_collapse(original_marks, 
#                                    "stromal_benign" = target_levels)
#     } else {
#       merged_marks <- original_marks # Fallback if neither exist
#     }
#     
#     # 3. Assign BOTH vectors to the ppp object as a data frame
#     marks(ppp_sample) <- data.frame(
#       original_type = original_marks,
#       merged_type   = merged_marks
#     )
#     saveRDS(ppp_sample, file_path)
# })
```

# Read the ppp objects

```{r}
sample_names <- readRDS(file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/TCGA_dataset/samples_name.rds")
path_dir = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/TCGA_dataset/samples_ppp"
ppp_list <- purrr::map(sample_names, \(sample_name){
    file_path <- file.path(path_dir, paste0(sample_name, ".rds"))
    ppp_sample <- readRDS(file_path)
    ppp_sample
})
names(ppp_list) <- sample_names
head(ppp_list)
```


Plot to check if the mask is correct and the points are inside
```{r eval=FALSE, include=FALSE}
poinside_plot <- \(ppp){
  ggplot() +
    # 1. Plot the tissue mask (Extracts pixels on the fly)
    geom_raster(data = as.data.frame(Window(ppp)),
                aes(x = x, y = y), fill = "grey80") +

    # 2. Plot the valid cells (Extracts coordinates on the fly)
    geom_point(data = as.data.frame(ppp),
               aes(x = x, y = y), color = "blue", size = 0.1, alpha = 0.5) +

    # 3. Plot the rejected cells (Extracts rejects on the fly)
    geom_point(data = as.data.frame(attr(ppp, "rejects")),
               aes(x = x, y = y), color = "red", size = 0.1, alpha = 0.5) +

    # Force standard Cartesian proportions so the image isn't squished
    coord_fixed() +
    theme_minimal() +
    labs(
      title = "Cell Alignment Check",
      subtitle = "Grey: Tissue Mask | Blue: Valid Cells | Red: Rejected Cells",
      x = "X Coordinate",
      y = "Y Coordinate"
    )
}
map(ppp_list, poinside_plot)
```

# Second order statistic analysis

npoints for each cell type and potential RAM cost
```{r eval=FALSE, include=FALSE}
# Define a safe threshold based on your RAM (e.g., 100 million pairs is usually safe for 16GB RAM)
# You might need to lower this to 50,000,000 if you have less RAM or other programs open.
MAX_PAIRS_THRESHOLD <- 100 * 10^6 

cost_npoints <- map(ppp_list, function(ppp_obj) {
  
  # 1. Extract the specific points
  n_neoplastic <- npoints(subset(ppp_obj, marks == "neoplastic"))
  n_stromal    <- npoints(subset(ppp_obj, marks == "stromal"))
  
  # 2. Calculate the "cost"
  pair_cost <- as.numeric(n_neoplastic) * as.numeric(n_stromal)
  
  # 3. Check if it exceeds the safety limit
  if (pair_cost > MAX_PAIRS_THRESHOLD) {
    message(sprintf("Skipping tile: Too large (%s pairs)", format(pair_cost, scientific = FALSE)))
    return(NULL) # Or return NA, or a placeholder
  }
  
  # 4. If safe, run the function
  tryCatch({
    Lcross(ppp_obj, i = "neoplastic", j = "stromal", correction = "best")
  }, error = function(e) {
    message("Error in Lcross: ", e$message)
    return(NULL)
  })
})
```

# memory check

Check if Lcross will crash - all unsupervised copilot code 
```{r eval=FALSE, include=FALSE}
library(scales) # For readable formatting (e.g., "1.2 GB")

estimate_lcross_memory <- function(ppp_obj, type_i, type_j, available_ram_gb = 60) {
  
  # 1. Check if marks exist and contain the requested types
  if (is.null(marks(ppp_obj))) {
    warning("Object has no marks.")
    return(NULL)
  }
  
  # 2. Count points
  # We use sum(marks == ...) which is faster than subsetting the whole object
  n_i <- sum(marks(ppp_obj) == type_i)
  n_j <- sum(marks(ppp_obj) == type_j)
  
  # 3. Calculate total pairs
  # We use as.numeric to prevent integer overflow for very large numbers
  n_pairs <- as.numeric(n_i) * as.numeric(n_j)
  
  # 4. Estimate Memory Usage
  # Base R double precision = 8 bytes per number.
  # Spatstat's Lcross/Kcross often needs to store:
  # - The distances (8 bytes)
  # - Edge correction weights (8 bytes)
  # - Temporary sorting/index vectors (4-8 bytes)
  # - Overhead for function calls
  # We use a safety multiplier of ~32 bytes per pair to be conservative.
  bytes_per_pair <- 32 
  est_bytes <- n_pairs * bytes_per_pair
  est_gb <- est_bytes / (1024^3)
  
  # 5. Determine if it's safe for your server
  # We leave 10% buffer for the OS and R overhead
  safe_limit <- available_ram_gb * 0.90 
  is_safe <- est_gb < safe_limit
  
  # 6. Return a summary list
  return(list(
    n_i = n_i,
    n_j = n_j,
    n_pairs = n_pairs,
    est_memory_gb = round(est_gb, 4),
    safe_to_run = is_safe,
    message = sprintf(
      "Tile Analysis: %s vs %s\n  - Count %s: %d\n  - Count %s: %d\n  - Pairs: %s\n  - Est. RAM: %s GB (Limit: %.1f GB) -> %s",
      type_i, type_j, type_i, n_i, type_j, n_j, 
      label_number(scale_cut = cut_short_scale())(n_pairs), 
      round(est_gb, 2), safe_limit,
      ifelse(is_safe, "SAFE", "DANGER")
    )
  ))
}
```

```{r eval=FALSE, include=FALSE}
# Example usage iterating over your list
# 1. Create a data frame of estimates
memory_check <- imap(ppp_list, function(ppp, name) {
  
  est <- estimate_lcross_memory(ppp, 
                                type_i = "neoplastic", 
                                type_j = "stromal", 
                                available_ram_gb = 60)
  
  # Return a tidy row for a dataframe
  data.frame(
    Sample = name,
    N_Neoplastic = est$n_i,
    N_Stromal = est$n_j,
    Pairs = est$n_pairs,
    Est_RAM_GB = est$est_memory_gb,
    Safe = est$safe_to_run
  )
})

# 2. Combine into a summary table
df_memory <- do.call(rbind, memory_check)

# 3. View the dangerous tiles
print(df_memory)

# 4. Run Lcross ONLY on safe tiles
# safe_tiles <- df_memory$Sample[df_memory$Safe == TRUE]
# 
# Lcross_results <- lapply(safe_tiles, function(tile_name) {
#   message("Processing: ", tile_name)
#   Lcross(ppp_sub[[tile_name]], i = "neoplastic", j = "stromal", correction = "best")
# })
```

Check memory usage based on rmax - all unsupervised copilot code 
```{r eval=FALSE, include=FALSE}
check_memory_rmax <- function(ppp_obj, type_i, type_j, rmax, available_gb = 60) {
  
  # 1. Get counts
  n_i <- npoints(subset(ppp_obj, marks == type_i))
  n_j <- npoints(subset(ppp_obj, marks == type_j))
  
  # 2. Get Area of the window
  area_val <- area(Window(ppp_obj))
  
  # 3. Calculate Global Density of type J (points per unit area)
  lambda_j <- n_j / area_val
  
  # 4. Estimate neighbors per point
  # Expected number of J points within rmax of an I point (assuming homogeneity)
  # Area of search circle = pi * r^2
  search_area <- pi * rmax^2
  avg_neighbors <- lambda_j * search_area
  
  # 5. Total estimated pairs to store
  # If points are highly clustered, this might be an underestimate, so we multiply by a safety factor (e.g., 2x or 3x)
  clustering_safety_factor <- 3 
  total_pairs_est <- n_i * avg_neighbors * clustering_safety_factor
  
  # 6. Convert to Memory (GB)
  # 32 bytes per pair is a conservative estimate for R vectors + overhead
  bytes_per_pair <- 32 
  mem_est_bytes <- total_pairs_est * bytes_per_pair
  mem_est_gb <- mem_est_bytes / 10^9
  
  # 7. Output results
  cat(sprintf("--- Memory Check for rmax = %s ---\n", rmax))
  cat(sprintf("Tile: %s x %s\n", type_i, type_j))
  cat(sprintf("Points: %s (i) vs %s (j)\n", n_i, n_j))
  cat(sprintf("Search Area (pi*r^2): %.2f\n", search_area))
  cat(sprintf("Est. Neighbors per point: %.2f (Global Avg)\n", avg_neighbors))
  cat(sprintf("Est. Total Pairs (with %dx cluster safety): %s\n", clustering_safety_factor, format(round(total_pairs_est), big.mark=",")))
  cat(sprintf("Est. RAM Usage: %.4f GB\n", mem_est_gb))
  
  if(mem_est_gb > (available_gb * 0.8)) {
    cat("⚠️  WARNING: MIGHT CRASH SERVER\n")
    return(FALSE)
  } else {
    cat("✅  SAFE TO RUN\n")
    return(TRUE)
  }
}

# --- Usage Example ---
map(ppp_list, \(ppp) {check_memory_rmax(ppp, "neoplastic", "stromal", rmax = 100)})
```


# Homogeneus ppp

## Lfunction estimate

L function between neoplastic and stromal
```{r}
# Lcross_homo <- map(ppp_list, \(ppp_obj) {
#      # Calculate Lcross
#      L <- Lcross(ppp_obj, i = "neoplastic", j = "stromal", correction = "border", rmax = 50)
#      L 
# }) 
```

Downsample to avoid crash

Function to estimate Lcross function for homogeneus point pattern between 2 marks and with downsample
```{r}
Lcross_personal <- function(ppp_obj, mark_i, mark_j, mark_column, downsample = TRUE, n_down = 10000, ...) {
  
  # 0. If marks are a dataframe, extract the specific column we want to use!
  if (is.data.frame(marks(ppp_obj))) {
    marks(ppp_obj) <- marks(ppp_obj)[[mark_column]]
  }
  # 1. Get the indices of the specific cell types dynamically
  idx_i <- which(marks(ppp_obj) == mark_i)
  idx_j <- which(marks(ppp_obj) == mark_j)
  
  # Safety check: ensure both cell types actually exist in the image!
  if(length(idx_i) < 2 | length(idx_j) < 2) {
    warning(paste("Not enough points for", mark_i, "and/or", mark_j, "to compute Lcross. Returning NULL."))
    return(NULL)
  }
  
  # 2. Handle Downsampling
  if (downsample) {
    # Randomly sample a maximum of n_down cells of each type
    keep_i <- sample(idx_i, min(length(idx_i), n_down))
    keep_j <- sample(idx_j, min(length(idx_j), n_down))
    
    # Subset the ppp object
    ppp_sub <- ppp_obj[c(keep_i, keep_j)]
    
  } else {
    # Even if we don't downsample, subsetting to JUST the two marks of interest 
    # saves a massive amount of memory for spatstat calculation
    ppp_sub <- ppp_obj[c(idx_i, idx_j)]
  }
  
  # 3. Calculate Lcross dynamically using the provided mark names
  L <- Lcross(ppp_sub, 
              i = mark_i, 
              j = mark_j, ...)
  
  return(L)
}
```

We investigate about the relationship between neoplastic nuclei and stromal/benign epitelium nuclei
```{r}

r_values <- seq(0, 100, length.out = 200)

Lcross_stroben <- map(ppp_list, \(ppp_obj) {
  Lcross_personal(ppp_obj, 
                  mark_i = "neoplastic", 
                  mark_j = "stromal_benign", 
                  mark_column = "merged_type", 
                  downsample = TRUE, n_down = 40000, rmax = 100, 
                  correction = "border", r = r_values)
})

#saveRDS(Lcross_stroben, file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/Lcross_stromal_benignEpitelial_91_samples.rds")
```

Let's check for NA in the result
```{r}
na_summary_per_sample <- imap_dfr(Lcross_stroben, \(fv_obj, sample_name) {
  
  # Handle cases where the Lcross calculation returned NULL
  if (is.null(fv_obj)) {
    return(tibble(
      sample = sample_name,
      r_na = NA_integer_,
      theo_na = NA_integer_,
      border_na = NA_integer_,
      status = "Skipped (NULL)"
    ))
  }
  
  # Convert the fv object to a standard dataframe
  df <- as.data.frame(fv_obj)
  
  # Count NAs for each specific column
  tibble(
    sample = sample_name,
    r_na = sum(is.na(df$r)),
    theo_na = sum(is.na(df$theo)),
    border_na = sum(is.na(df$border)),
    status = "Computed"
  )
})

# View the per-sample breakdown
print(na_summary_per_sample)

# 2. Count NAs for EVERYTHING (Grand Totals)
grand_totals <- na_summary_per_sample |> 
  summarise(
    total_r_NAs = sum(r_na, na.rm = TRUE),
    total_theo_NAs = sum(theo_na, na.rm = TRUE),
    total_border_NAs = sum(border_na, na.rm = TRUE),
    total_failed_samples = sum(status == "Skipped (NULL)")
  )

# View the total counts
print(grand_totals)
```

There should be some NA for overflow, but there is no one... ok i guess

Plot the results

```{r}
plot_fv_list <- function(list_of_funcs, comparison_df, correction = "border") {
  
  # 1. Run the map_dfr block dynamically on whatever list you provide
  df_lcross <- map_dfr(names(list_of_funcs), function(slide_name) {
    
    # Extract the specific L-function object from the provided list
    L_obj <- list_of_funcs[[slide_name]]
    
    as.data.frame(L_obj) |> 
      mutate(
        Slide = slide_name,
        # Use .data[[correction]] to dynamically grab the 'border' column (or 'iso', etc.)
        L_r_minus_r = .data[[correction]] - r,  
        
        # Extract the marks names exactly as you designed
        mark_i = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][1]),
        mark_j = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][2])
      )
  })
  
  # 2. Join the metadata (comparison_benign) into the dataframe
  df_lcross <- df_lcross |> 
    mutate(
      # Extract the first 12 characters to match "TCGA-XX-XXXX"
      patientID = substr(Slide, 1, 12)
    ) |> 
    # Merge the cluster info based on the patientID
    left_join(comparison_df, by = "patientID") |> 
    # Convert cluster_ilaria to a factor so ggplot uses discrete colors instead of a continuous gradient
    mutate(cluster_ilaria = as.factor(cluster))
  
  # 3. Build the ggplot
  # Note: group = Slide ensures each patient gets their own line
  #       color = cluster_ilaria applies the cluster colors
  p <- ggplot(df_lcross, aes(x = r, y = L_r_minus_r, group = Slide, color = cluster_ilaria)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(alpha = 0.7, linewidth = 0.6) + 
    theme_minimal() +
    labs(
      title = paste0("L-cross Function: ", df_lcross$mark_i[1], " vs ", df_lcross$mark_j[1]),
      subtitle = "Centered L-function (L(r) - r). Positive = Attraction, Negative = Repulsion",
      x = "Radius (r)",
      y = "L(r) - r",
      color = "Cluster" # Renamed legend title
    ) +
    theme(legend.position = "right")
  
  return(p)
}
```

load the comparison_df
```{r}
meta_ilaria <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/TCGA_dataset/TCGA_data/survival_dataset_&_metadata.rds")
```


```{r}
plot_colored <- plot_fv_list(
  list_of_funcs = Lcross_stroben, 
  comparison_df = meta_ilaria, 
  correction = "border"
)

plot_colored
```

# Functional Data Analysis


# generic function for smoothing funs
```{r}
fsmooth <- function(fcross_list, fun, M = 6, genLfun.fd = 4, centrata = F, ...)
{

  # 1. Calculate raw L-functions (Common for both)
  r <- fcross_list[[1]]$r
  mat_fun <- sapply(fcross_list, function(x) x$border)
  
  # 2. Apply centering ONLY if requested
  if(centrata == TRUE){
    mat_fun <- sweep(mat_fun, MARGIN = 1, STATS = r, FUN = "-")
  }

  # 3. Create Basis and Smooth
  K <- length(r) + M - 2
  genbasis <- create.bspline.basis(range(r), K, M, r)
  genfdPar <- fdPar(genbasis, genLfun.fd, lambda = 1e-11)
  
  genfdSmooth <- smooth.basis(r, mat_fun, genfdPar)
  Genfun.fd <- genfdSmooth$fd
  
  # 4. Set Names
  fdn_dist <- paste("Distanza (da 0 a", max(r), "micron)")
  fdnames <- list(fdn_dist,
                  "Gene" = colnames(mat_fun), # Crucial: use names of surviving columns
                  "Funzione")
  Genfun.fd$fdnames <- fdnames
  
  return(list(fd = Genfun.fd, fdPar = genfdPar, 
              basis = genbasis, mat = mat_fun, r = r))
}
```



```{r}
Lfun_stroben <- fsmooth(Lcross_stroben, centrata = T)
```

```{r}
plot(Lfun_stroben$fd)
lines(c(0,100), c(0,0), col = "gold", lwd = 2)
```

## Comparison between neoplastic and benign.epithelial

### Functional PCA

Calcoliamo fpca per 4 componenti e mostriamo il risultato sulla rotazione VARIMAX

```{r}
fpca <- pca.fd(Lfun_stroben$fd, nharm = 4, Lfun_stroben$fdPar)
rownames(fpca$scores) <- names(Lcross_stroben) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca)
```

### fPCA plot ruotate per VARIMAX

The plots below show the mean function and the effects of adding (+) and subtracting (−) a
suitable multiple of the first (resp. second) principal component function.

```{r eval=FALSE, include=FALSE}
plot.pca.fd(fpcaVarmx, cex.main=0.9)
```

## Clustering

Applichiamo due metodi di clustering sugli scores delle prime 4 componenti.
Clustering gerarchico con metodo Ward e kmeans

```{r}
set.seed(123)
fpca_km <- kmeans(fpca$scores[,c(1,4)], 4)
fpca_hc <- hclust(dist(fpca$scores[,c(1,4)]), method = "ward.D2")
memb <- cutree(fpca_hc, k = 4)
```

Grafico del dendogramma per il clustering gerarchico, scegliamo di dividere in 5 cluster.

```{r}
dhc <- as.dendrogram(fpca_hc)
plot(dhc, cex = 0.1, horiz = T)
#ggdendrogram(dhc, rotate = TRUE, size = 2)
table(kmean = fpca_km$cluster)
table(ward.d2 = memb)
```

```{r}
plot(fpca$scores[,1], fpca$scores[,2], col = fpca_km$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca$scores[,1], fpca$scores[,2], col = memb, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

## Check the outlier
```{r}
poinside_plot <- \(ppp, title_p = "Spatial point pattern"){
  ggplot() +
    # 1. Plot the tissue mask (Extracts pixels on the fly)
    geom_raster(data = as.data.frame(Window(ppp)), 
                aes(x = x, y = y), fill = "grey80") +
    
    # 2. Plot the valid cells (Extracts coordinates on the fly)
    geom_point(data = as.data.frame(ppp), 
               aes(x = x, y = y, color = merged_type), size = 0.05, alpha = 0.5) +

    # Force standard Cartesian proportions so the image isn't squished
    coord_fixed() +
    theme_minimal() +
    labs(
      title = title_p,
      subtitle = "Colored by cell type with merged types",
      x = "X Coordinate", 
      y = "Y Coordinate"
    )
}
```

Compare it with the patientID ppp with the lowest PC1 and the PC1 nearest to zero

```{r}
which.max(fpca$scores[,1]) # TCGA-23-1121-01Z-00-DX1.E2F25441-32C3-46BF-A845-CB4FA787E8CB OUTLIER
which.min(fpca$scores[,1]) # TCGA-23-1030-01Z-00-DX1.FC93147F-FB55-4C62-B187-337A5C107096 
which((fpca$score[,1] > -5) & (fpca$score[,1] < 5)) # TCGA-57-1582-01Z-00-DX1.BDF02DAA-E520-46BD-A479-58CAA4354B1A the 83rd row

poinside_plot(ppp_list[[24]], title_p = "highest pc1 and pc2")
poinside_plot(ppp_list[[which.min(fpca$scores[,1])]], title_p = "lowest pc1 highest pc2")
poinside_plot(ppp_list[[83]], title_p = "near zero pc1 and pc2")
```

let's remove the sample 24 and redo the analysis

```{r}
df_ilaria <- meta_ilaria |> 
  dplyr::select(patientID, cluster)
# 3. Clean Dataset B to match IDs
# We extract the first 12 characters (e.g., "TCGA-25-1326")
df_clust <- data.frame(
  patientID = substr(names(memb), 1, 12),
  cluster_kmean = fpca_km$cluster,
  cluster_ward = memb
) |> print()

# 4. Merge to find correspondence
comparison_stroben <- inner_join(df_ilaria, df_clust, by = "patientID") |> 
  arrange(cluster_ward, patientID)

# 5. Display the Result
print(comparison_stroben)

table("ilaria_clust" = comparison_stroben$cluster, "valerio_clust" = comparison_stroben$cluster_ward)
```

## Comparison between results

```{r eval=FALSE, include=FALSE}
df_A <- data.frame(
  patientID = c(
    "TCGA-OY-A56P", "TCGA-23-1021", "TCGA-23-2072", "TCGA-23-1031", "TCGA-23-1117", # Cluster 1
    "TCGA-25-2401", "TCGA-25-2399", "TCGA-25-1625", "TCGA-25-1328", "TCGA-25-1871", # Cluster 2
    "TCGA-25-1315", "TCGA-25-1326", "TCGA-25-2391", "TCGA-25-1635", "TCGA-25-1627", # Cluster 3
    "TCGA-57-1582", "TCGA-13-A5FU", "TCGA-57-1993", "TCGA-57-1994", "TCGA-57-1583"  # Cluster 4
  ),
  cluster_ilaria = c(
    rep(1, 5), 
    rep(2, 5), 
    rep(3, 5), 
    rep(4, 5)
  )
)

# 3. Clean Dataset B to match IDs
# We extract the first 12 characters (e.g., "TCGA-25-1326")
df_clust <- data.frame(
  patientID = substr(names(vec_kmean), 1, 12),
  cluster_kmean = fpca_km$cluster,
  cluster_ward = memb
)

# 4. Merge to find correspondence
comparison_stroma <- inner_join(df_A, df_clust, by = "patientID") |> 
  arrange(cluster_kmean, patientID)

# 5. Display the Result
#print(comparison_stroma)

table("ilaria_clust" = comparison_stroma$cluster_ilaria, "valerio_clust" = comparison_stroma$cluster_kmean)
```

Comparison between the 2 results of my work
```{r eval=FALSE, include=FALSE}
table("benign_clust" = comparison_benign$cluster_kmean, "stroma_cluster" = comparison_stroma$cluster_kmean)
```

# Survival analysis

```{r}
ov_meta <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/TCGA_dataset/TCGA_data/survival_dataset_&_metadata.rds")
```


```{r}
df_combined <- ov_meta |> 
  # Select only the columns we want to add from the first df
  inner_join(df_clust |> dplyr::select(patientID, cluster_kmean), by = "patientID")
```

```{r}
set.seed(42)
fit <- survfit(Surv(time, vital_status) ~ cluster, data = df_combined)
(km_plot <- ggsurvplot(fit, data = df_combined, risk.table = TRUE, pval = TRUE, xlim = c(0,3000)))
pairwise_survdiff(Surv(time, vital_status) ~ cluster, data = df_combined)
```

# REMOVE THE OUTLIER

```{r}
Lcross_stroben_filtered <- Lcross_stroben[names(Lcross_stroben) != "TCGA-23-1121-01Z-00-DX1.E2F25441-32C3-46BF-A845-CB4FA787E8CB"]
```

```{r}
Lfun_stroben_filtered <- fsmooth(Lcross_stroben_filtered, centrata = T)
plot(Lfun_stroben_filtered$fd)
lines(c(0,100), c(0,0), col = "gold", lwd = 2)
```

```{r}
fpca_filt <- pca.fd(Lfun_stroben_filtered$fd, nharm = 4, Lfun_stroben_filtered$fdPar)
rownames(fpca_filt$scores) <- names(Lcross_stroben_filtered) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca_filt)


set.seed(123)
fpca_km_filt <- kmeans(fpca_filt$scores[,c(1,4)], 4)
fpca_hc_filt <- hclust(dist(fpca_filt$scores[,c(1,4)]), method = "ward.D2")
memb_filt <- cutree(fpca_hc_filt, k = 4)
```

```{r}
#ggdendrogram(dhc, rotate = TRUE, size = 2)
table(kmean = fpca_km_filt$cluster)
table(ward.d2 = memb_filt)
```

```{r}
plot(fpca_filt$scores[,1], fpca_filt$scores[,2], col = fpca_km_filt$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca_filt$scores[,1], fpca_filt$scores[,2], col = memb_filt, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

```{r}
df_ilaria <- meta_ilaria |> 
  dplyr::select(patientID, cluster)
# 3. Clean Dataset B to match IDs
# We extract the first 12 characters (e.g., "TCGA-25-1326")
df_clust_filt <- data.frame(
  patientID = substr(names(memb_filt), 1, 12),
  cluster_kmean = fpca_km_filt$cluster,
  cluster_ward = memb_filt
) |> print()

# 4. Merge to find correspondence
comparison_stroben_filt <- inner_join(df_ilaria, df_clust_filt, by = "patientID") |> 
  arrange(cluster_ward, patientID)

# 5. Display the Result
print(comparison_stroben_filt)

table("ilaria_clust" = comparison_stroben_filt$cluster, "valerio_clust" = comparison_stroben_filt$cluster_ward)
```

# Survival analysis

```{r}
ov_meta <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/TCGA_dataset/TCGA_data/survival_dataset_&_metadata.rds")
```


```{r}
df_combined_filt <- ov_meta |> 
  # Select only the columns we want to add from the first df
  inner_join(df_clust_filt |> dplyr::select(patientID, cluster_kmean, cluster_ward), by = "patientID")
```

```{r}
set.seed(42)
fit_filt <- survfit(Surv(time, vital_status) ~ cluster_kmean, data = df_combined_filt)
(km_plot_filt <- ggsurvplot(fit_filt, data = df_combined_filt, risk.table = TRUE, pval = TRUE, xlim = c(0,3000)))
pairwise_survdiff(Surv(time, vital_status) ~ cluster_kmean, data = df_combined_filt)
```

# Sketching

```{r}
sketch_spatial_fast <- function(ppp_obj, idx, n_target) {
  if (length(idx) <= n_target) return(idx)
  
  # Corrected math: Number of bins per axis should be the square root 
  # of the total desired boxes (n_target * 5)
  grid_res <- ceiling(sqrt(n_target * 5))
  
  # Get coordinates
  x_coords <- ppp_obj$x[idx]
  y_coords <- ppp_obj$y[idx]
  
  # Assign each point to an X and Y grid bin
  x_bin <- findInterval(x_coords, seq(min(x_coords), max(x_coords), length.out = grid_res))
  y_bin <- findInterval(y_coords, seq(min(y_coords), max(y_coords), length.out = grid_res))
  
  # Create a unique ID for each grid square
  grid_id <- paste(x_bin, y_bin, sep = "_")
  
  # Shuffle the data randomly
  shuffled_order <- sample(length(idx))
  idx_shuffled <- idx[shuffled_order]
  grid_id_shuffled <- grid_id[shuffled_order]
  
  # By taking the first occurrence of each grid_id, we get exactly 1 random cell per grid square!
  kept_idx <- idx_shuffled[!duplicated(grid_id_shuffled)]
  
  # If the grid gave us more points than n_target, randomly downsample to exactly n_target
  if (length(kept_idx) > n_target) {
    kept_idx <- sample(kept_idx, n_target)
  }
  
  return(kept_idx)
}
```

```{r}
# Pure R Spatial Sketching Function
sketch_spatial_R <- function(ppp_obj, idx, n_target) {
  if (length(idx) <= n_target) return(idx)
  
  # Extract X and Y coordinates for these specific points
  coords <- cbind(x = ppp_obj$x[idx], y = ppp_obj$y[idx])
  
  # Cluster the physical space into 'n_target' even regions
  # Using MacQueen algorithm because it is blazingly fast for 2D coordinates
  km <- kmeans(coords, centers = n_target, algorithm = "MacQueen", iter.max = 15)
  
  # Pick one random cell from each spatial cluster
  # This forces the downsampled cells to perfectly trace the tissue structure
  sketched_rel_idx <- tapply(seq_along(idx), km$cluster, function(x) sample(x, 1))
  
  return(idx[sketched_rel_idx])
}
```


```{r}
Lcross_personal <- function(ppp_obj, mark_i, mark_j, mark_column, downsample = TRUE, downsample_method = c("sketch_R", "sketch_fast", "standard"), n_down = 10000, ...) {
  
  # Ensure a valid method is selected (defaults to "sketch_R" if none is specified)
  downsample_method <- match.arg(downsample_method)
  
  # 0. If marks are a dataframe, extract the specific column we want to use
  if (is.data.frame(marks(ppp_obj))) {
    marks(ppp_obj) <- marks(ppp_obj)[[mark_column]]
  }
  
  # 1. Get the indices of the specific cell types dynamically
  idx_i <- which(marks(ppp_obj) == mark_i)
  idx_j <- which(marks(ppp_obj) == mark_j)
  
  if(length(idx_i) < 2 | length(idx_j) < 2) {
    warning(paste("Not enough points for", mark_i, "and/or", mark_j, "to compute Lcross. Returning NULL."))
    return(NULL)
  }
  
  # 2. Handle Downsampling
  if (downsample) {
    if (downsample_method == "sketch_R") {
      # MacQueen k-means sketching
      keep_i <- sketch_spatial_R(ppp_obj, idx_i, n_down)
      keep_j <- sketch_spatial_R(ppp_obj, idx_j, n_down)
      
    } else if (downsample_method == "sketch_fast") {
      # Fast grid-based sketching (requires ppp_obj to be passed)
      keep_i <- sketch_spatial_fast(ppp_obj, idx_i, n_down)
      keep_j <- sketch_spatial_fast(ppp_obj, idx_j, n_down)
      
    } else if (downsample_method == "standard") {
      # Standard random downsampling
      keep_i <- sample(idx_i, min(length(idx_i), n_down))
      keep_j <- sample(idx_j, min(length(idx_j), n_down))
    }
    
    # Subset the ppp object
    ppp_sub <- ppp_obj[c(keep_i, keep_j)]
    
  } else {
    ppp_sub <- ppp_obj[c(idx_i, idx_j)]
  }
  
  # 3. Calculate Lcross
  L <- Lcross(ppp_sub, i = mark_i, j = mark_j, ...)
  
  return(L)
}
```

```{r}
Lcross_sketch <- map(ppp_list, \(ppp_obj) {
  Lcross_personal(ppp_obj, 
                  mark_i = "neoplastic", 
                  mark_j = "stromal_benign", 
                  mark_column = "merged_type", 
                  downsample = TRUE, 
                  downsample_method = "sketch_fast",
                  n_down = 10000, rmax = 100, 
                  correction = "border", r = r_values)
})
Lcross_sketch <- Lcross_sketch[names(Lcross_sketch) != "TCGA-23-1121-01Z-00-DX1.E2F25441-32C3-46BF-A845-CB4FA787E8CB"]
```

```{r}
plot_colored <- plot_fv_list(
  list_of_funcs = Lcross_sketch, 
  comparison_df = meta_ilaria, 
  correction = "border"
)

plot_colored
```

### Remove the outliner

```{r}
Lcross_sketch <- Lcross_sketch[names(Lcross_sketch) != "TCGA-23-1121-01Z-00-DX1.E2F25441-32C3-46BF-A845-CB4FA787E8CB"]
```


```{r}
Lfun_sketch <- fsmooth(Lcross_sketch, centrata = T)
```

```{r}
plot(Lfun_sketch$fd)
lines(c(0,100), c(0,0), col = "gold", lwd = 2)
```

```{r}
fpca_sk <- pca.fd(Lfun_sketch$fd, nharm = 4, Lfun_sketch$fdPar)
rownames(fpca_sk$scores) <- names(Lcross_sketch) # "nome" dei pazienti (numero del paziente)

set.seed(123)
fpca_km_sk <- kmeans(fpca_sk$scores[,c(1,4)], 4)
fpca_hc_sk <- hclust(dist(fpca_sk$scores[,c(1,4)]), method = "ward.D2")
memb_sk <- cutree(fpca_hc_sk, k = 4)

table(kmean = fpca_km_sk$cluster)
table(ward.d2 = memb_sk)


```

```{r}
plot(fpca_sk$scores[,1], fpca_sk$scores[,2], col = fpca_km_sk$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca_sk$scores[,1], fpca_sk$scores[,2], col = memb_sk, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

```{r}
df_ilaria <- meta_ilaria |> 
  dplyr::select(patientID, cluster)
# 3. Clean Dataset B to match IDs
# We extract the first 12 characters (e.g., "TCGA-25-1326")
df_clust_sketch <- data.frame(
  patientID = substr(names(memb_sk), 1, 12),
  cluster_kmean = fpca_km_sk$cluster,
  cluster_ward = memb_sk
) |> print()

# 4. Merge to find correspondence
comparison_sketch <- inner_join(df_ilaria, df_clust_sketch, by = "patientID") |> 
  arrange(cluster_ward, patientID)

# 5. Display the Result
print(comparison_sketch)

table("ilaria_clust" = comparison_sketch$cluster, "valerio_ward" = comparison_sketch$cluster_ward)
table("ilaria_clust" = comparison_sketch$cluster, "valerio_kmean" = comparison_sketch$cluster_kmean)
```

## Survival analysis

```{r}
ov_meta <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/TCGA_data/survival_dataset_&_metadata.rds")
```


```{r}
df_combined_sketch <- ov_meta |> 
  # Select only the columns we want to add from the first df
  inner_join(df_clust_sketch |> dplyr::select(patientID, cluster_kmean), by = "patientID")
```

```{r}
set.seed(42)
fit <- survfit(Surv(time, vital_status) ~ cluster_kmean, data = df_combined_sketch)
(km_plot <- ggsurvplot(fit, data = df_combined_sketch, risk.table = TRUE, pval = TRUE, xlim = c(0,3000)))
pairwise_survdiff(Surv(time, vital_status) ~ cluster_kmean, data = df_combined_sketch)
```

COSE DA FARE:

dopo aver rimosso casi strani, prendo tutti i pazienti e faccio tutta la pipe
con i gruppi di clustering faccio survival. 

sketching: subsampling 50k cells, ribilianciare i punti con pesi

unisco benign e stromal
match i neoplastic con le altre.