---
title: "Figures for the SIS's short paper"
author: "Valerio Reffo"
date: "2026-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(anndataR)
library(ggspavis)
library(patchwork)
library(spatstat.geom)
library(spatstat)
library(sosta)
library(purrr)
library(readxl)
library(dplyr)
library(spatialFDA)
library(fda)
# survival analysis
library(survival)
library(survminer)

#palette 
library(paletteer)
```

Where to save them
```{r}
plot_folder <- "~/repositories/spatialPointProcess_clustering/figures/to_publish"
```

Data
```{r}
spe <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/figures_sample.rds")
spe$type <- forcats::fct_reorder(spe_type, )
```

# First figure

Thumbnails of the original image and the ppp

Function for the spatial plot
```{r}
spatial_coords_plot <- function(spe, cell_type_key = "type", ...) {
  # data
  coords <- spatialCoords(spe)
  plot_df <- data.frame(
    x = coords[, 1],
    y = coords[, 2],
    cell_type = colData(spe)[[cell_type_key]]
  )
  # plot
  plot <- ggplot(plot_df, aes(x = x, y = y, colour = cell_type)) +
    geom_point(...) +
    coord_fixed() +
    theme_void() +
    #scale_color_manual(values = color_vector) +
    labs(color = "Cell types") + # Set the legend title to the column name
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    ) +
    # This override.aes increases the size of the points inside the legend
    guides(color = guide_legend(override.aes = list(size = 5, alpha = 1)),
           ) +
    scale_y_reverse()

  return(plot)
}

```

Palette candidate
scale_colour_paletteer_d("tvthemes::CrazyLaceAgate")
scale_fill_paletteer_d("tvthemes::CrazyLaceAgate")

scale_colour_paletteer_d("feathers::bee_eater")
#00346EFF, #007CBFFF, #06ABDFFF, #EDD03EFF, #F5A200FF, #6D8600FF, #424D0CFF

```{r}
color_vec <- c("benign epithelial" = "#00346EFF",
"inflammatory" = "#6D8600FF",
"necrotic" = "#F5A200FF",
"neoplastic" = "#EDD03EFF",
"no label" = "#D3D3D3FF",
"stromal" = "#007CBFFF"
)
legend_order <- c("stromal", "benign epithelial", "inflammatory", "necrotic", "neoplastic", "no label")
```


```{r}
p1 <- plotVisium(prova, spots = FALSE)
p2 <- spatial_coords_plot(prova, size = 0.02, alpha = 0.4) +
  scale_color_manual(
    values = color_vec,
    breaks = legend_order
  )
plot1 <- p1 + p2

ggsave(filename = "/figure1.pdf", plot = plot1, path = plot_folder, dpi = 320, width = 10)
```


# Figure 2

2 panels, one with Lcross mean curves for each cluster, one with scores of pc colored by cluster

Data
```{r}
Lcross_list <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/Lcross_stromal_benignEpitelial_91_samples.rds")
Lcross_filtered <- Lcross_list[names(Lcross_list) != "TCGA-23-1121-01Z-00-DX1.E2F25441-32C3-46BF-A845-CB4FA787E8CB"]
```

```{r}
Lfun_stroben <- fsmooth(Lcross_stroben_filtered, centrata = T)
```


```{r}
# Creates a named list of mean fd objects
mean_fd_list <- unique(df_clust_filt$cluster_kmean) |> 
  set_names(paste0("Cluster_", unique(df_clust_filt$cluster_kmean))) |> 
  map(\(k) {
    # Subset the fd object for this specific cluster and calculate the mean
    mean.fd(Lfun_stroben$fd[which(df_clust_filt$cluster_kmean == k)])
  })

# Access cluster 1's mean function:
#plot(mean_fd_list$Cluster_1)
```

color palette
```{r}
color_vec_4 <- c(
  "Cluster_1" = '#E27B0CFF',
  "Cluster_2" = '#FED105FF',
  "Cluster_3" = '#6CA167FF',
  "Cluster_4" = '#145A76FF'
)

```

Panel 1
```{r}
library(fda)
library(dplyr)
library(purrr)
library(ggplot2)
library(ggthemes)

# 1. Define a grid of points to evaluate the functions on.
# We extract the range directly from the basis of the first mean function
basis_range <- mean_fd_list[[1]]$basis$rangeval
eval_grid <- seq(basis_range[1], basis_range[2], length.out = 100)

# 2. Evaluate each mean function and combine into a tidy dataframe
mean_plot_df <- imap_dfr(mean_fd_list, \(fd_obj, cluster_name) {
  tibble(
    r = eval_grid,
    # eval.fd returns a matrix, we coerce it to a vector
    value = as.vector(eval.fd(eval_grid, fd_obj)),
    cluster = cluster_name
  )
})

# 3. Plot with ggplot2
plot2_a <- ggplot(mean_plot_df, aes(x = r, y = value, color = cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") + # Added horizontal line
  geom_line(linewidth = 2.5) +
  #theme_par() +
  theme_few() +
  theme(
    legend.title = element_text(size = 36, face = "bold"),
    legend.text  = element_text(size = 26),
    axis.title = element_text(size = 26),
    axis.text  = element_text(size = 18),
  ) +
  labs(
    title = "",
    x = "r (micron)",
    y = "L(r) - r",
    color = "Clusters"
  ) +
  scale_color_manual(
    values = color_vec_4
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1), ncol = 1))

plot2_a

plot2_a <- ggplot(mean_plot_df, aes(x = r, y = value, color = cluster)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(linewidth = 1.5) + 
    theme_minimal() +
    labs(
      title = "",
      subtitle = "",
      x = "Radius (r)",
      y = "L(r) - r",
      color = "Sample Type"
    ) +
    scale_color_manual(
    values = color_vec_4
  ) +
    theme(legend.position = "right",
          legend.title = element_text(size = 36, face = "bold"),
          legend.text  = element_text(size = 26),
          axis.title = element_text(size = 26),
          axis.text  = element_text(size = 18),
          ) +
    guides(color = guide_legend(override.aes = list(size = 7, alpha = 1, linewidth = 4), ncol = 1))
  
plot2_a
```

panel 2 - scores fPCA
```{r}
fpca_plot_df <- tibble(
  FPC1 = fpca_filt$scores[, 1],
  FPC2 = fpca_filt$scores[, 2],
  Cluster = factor(paste0("Cluster_", fpca_km_filt$cluster))
)

# 3. Create the ggplot
plot2_b <- ggplot(fpca_plot_df, aes(x = FPC1, y = FPC2, color = Cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  geom_point(size = 2.5) + # shape = 3 is the equivalent of pch = 3
  scale_color_manual(values = color_vec_4) +
  theme_few() +
  labs(
    title = "",
    x = "fPC 1",
    y = "fPC 2"
  )
```

```{r}
fig2 <- plot2_a | plot2_b

ggsave(filename = "/figure2.pdf", plot = fig2, path = plot_folder, dpi = 320, width = 20)
```

Second try for figure 2
```{r}
# 1. Define plot2_a
plot2_a <- ggplot(mean_plot_df, aes(x = r, y = value, color = cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") + 
  geom_line(linewidth = 1) +
  theme_few() +
  labs(
    title = "",
    x = "r (micron)",
    y = "L(r) - r",
    color = "Cluster"
  ) +
  scale_color_manual(values = color_vec_4)
plot2_a <- plot2_a + theme(legend.position = "none")

# 2. Define plot2_b (Assuming it uses the same structure but different data/metrics)
plot2_b <- ggplot(fpca_plot_df, aes(x = FPC1, y = FPC2, color = Cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  geom_point(size = 2.5) + # shape = 3 is the equivalent of pch = 3
  scale_color_manual(values = color_vec_4) +
  theme_few() +
  labs(
    title = "",
    x = "fPC 1",
    y = "fPC 2"
  )

# 3. Combine with Patchwork and apply global formatting
fig2 <- (plot2_a | plot2_b) + 
  plot_layout(guides = "collect") & # Collects legends into 1 shared legend
  theme(
    # Make axis titles and tick texts larger
    axis.title = element_text(size = 16, face = "bold"),
    axis.text  = element_text(size = 14),
    
    # Make legend title and text larger
    legend.title = element_text(size = 16, face = "bold"),
    legend.text  = element_text(size = 14),
    
    # Increase the overall spacing/size of the legend keys
    legend.key.size = unit(2, "lines") 
  ) &
  # Make the actual color keys (lines/balls) inside the legend thicker/larger
  guides(color = guide_legend(override.aes = list(linewidth = 2, size = 5)))

# Print the final plot
fig2
ggsave(filename = "/figure2.pdf", plot = fig2, path = plot_folder, dpi = 320, width = 20)

```


# Figures 3

KM survival curves
```{r}
color_vec_4b <- c(
  "cluster_kmean=1" = '#E27B0CFF',
  "cluster_kmean=2" = '#FED105FF',
  "cluster_kmean=3" = '#6CA167FF',
  "cluster_kmean=4" = '#145A76FF'
)

set.seed(42)
fit <- survfit(Surv(time, vital_status) ~ cluster_kmean, data = df_combined_filt)
fig3 <- ggsurvplot(fit, data = df_combined_filt, 
                    palette = color_vec_4b, risk.table = FALSE, 
                   pval = TRUE, xlim = c(0,3000), linewidth = 3,
                   pval.size = 10)
fig3$plot <- fig3$plot +
  theme(
    legend.title = element_text(size = 36, face = "bold"),
    legend.text  = element_text(size = 26),
    axis.title = element_text(size = 26),
    axis.text  = element_text(size = 18),
  ) +
  guides(color = guide_legend(
    override.aes = list(linewidth = 3), 
    nrow = 2,       # <--- Forces the legend into exactly 2 lines (rows)
    byrow = TRUE    # <--- (Optional) Fills the legend left-to-right, then moves to the next line
  ))
fig3
```

```{r}
fig3_better <- fig3$plot / fig3$table  

fig3_final <- fig3_better + 
  plot_layout(heights = c(3, 1))
```

```{r}

ggsave(filename = "/figure3.pdf", plot = fig3_final, path = plot_folder, dpi = 300, width = 10, height = 8)
```










