---
title: "Different fPCA methods"
author: "Valerio Reffo"
date: "2026-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(fdapace)
library(SpatialExperiment)
library(spatstat)
library(dplyr)
library(purrr)
```

# Creation of the FPCA inputs object 

I need a vector with r values, the matrix of values of Lcross for each patient, and a patientID vector
```{r}
# 1. Iterate through the list and combine
df_long <- imap_dfr(Lcross_benign, function(obj, name) {
  
  # Coerce the 'fv' object to a standard data.frame
  df <- as.data.frame(obj)
  
  # Select columns and return
  data.frame(
    r = df$r,
    border = df$border - df$r,
    patientID = substr(name, 1, 12) # Extract TCGA-XX-XXXX
  )
})

# View the result
head(df_long)

pace_input <- MakeFPCAInputs(IDs = df_long$patientID, tVec = df_long$r, yVec = df_long$border)
```


```{r}
fpcaObj_benign <- FPCA(pace_input$Ly, pace_input$Lt, list(plot = TRUE, methodMuCovEst = 'smooth', userBwCov = 10))
```

Inspection of the 2 PC

```{r}
require('ks')

CreatePathPlot(fpcaObj_benign, K = 1, main = 'K = 1', pch = 4); grid()
CreatePathPlot(fpcaObj_benign, K = 2, main = 'K = 2', pch = 4) ; grid()
```

Outlier detection??

```{r}

CreateOutliersPlot(fpcaObj_benign, optns = list(K = 2, variant = 'KDE'))
CreateFuncBoxPlot(fpcaObj_benign, xlab = 'r', ylab = 'Lcross', optns = list(K =2, variant='bagplot'))
```

Rate of change of this function can be of interest in FDA - tangentially the derivative of Kcross (of which Lcross is a function sqrt(Kcross/r)) is a function of interest itself, the pair correlation function
```{r}

CreatePathPlot(fpcaObj_benign, K = 2, main = 'K = 2', showObs = FALSE) ; grid()
CreatePathPlot(fpcaObj_benign, K = 1, main = 'K = 1', showObs = FALSE, derOptns = list(p = 1, bw = 1.01 , kernelType = 'epan'), ylim = c(-1000, 1000) ) ; grid()
```

