---
title: "Ilaria Data - HoverNet segmentation"
author: "Valerio Reffo"
date: "2026-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(SpatialExperiment)
library(anndataR)
library(ggspavis)
library(patchwork)
library(spatstat.geom)
library(spatstat)
library(sosta)
library(purrr)
source("~/repositories/spatialPointProcess_clustering/helper_function.R", echo = TRUE)
```

```{r}
data_path <- '/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/h5ad/TCGA-WR-A838-01Z-00-DX1.5FE22DE4-CEFB-45F6-9299-505023A8F3BA.h5ad'
# prova <- read_h5ad(data_path, as = "SingleCellExperiment") perdiamo obsm (coordinate spaziali) quindi convertiamo manualmente
```

Function to transform anndata to SpatialExperiment

```{r}
ann2spe <- \(data_path, spat_coords_name = c("x", "y"), img_path = NULL, ...) {
  ann <- anndataR::read_h5ad(data_path)
  #object to build SpatialExperiment
  counts <- t(ann$X) # transpose it
  cd <- ann$obs
  meta <- ann$uns
  spat <- ann$obsm$spatial
  colnames(spat) <- spat_coords_name
  cd <- cbind(cd, spat)
  spe <- SpatialExperiment::SpatialExperiment(assays = base::list("counts" = counts), 
                                              colData = cd,
                                              metadata = meta,
                                              spatialCoordsNames = spat_coords_name,
                                              imageSources = img_path, ...)
  spe
}
```

```{r}
spe <- ann2spe(data_path, img_path = "/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/thumb/TCGA-WR-A838-01Z-00-DX1.5FE22DE4-CEFB-45F6-9299-505023A8F3BA.png")
spe
```

```{r}
#plotVisium(spe, spots = FALSE)
```

```{r}
#plotCoords(spe, annotate = "type", in_tissue = NULL, point_size = 0.1)
```

```{r}
plotVisium(spe, spots = FALSE) | 
  plotCoords(spe, annotate = "type", in_tissue = NULL, point_size = 0.1)
```

# Window

## ripras
```{r}
# 1. Extract the coordinates from your SpatialExperiment object
coords <- spatialCoords(spe)

window_polygon <- convexhull.xy(x = coords[, "x"], y = coords[, "y"])

plot(window_polygon, main = "Calculated Window Polygon")
print(window_polygon)
```

## Points intensity based window estimate
```{r}
n <- estimateReconstructionParametersSPE(
    spe,
    marks = "type",
    imageCol = "sample_id",
    markSelect = NULL,
    plotHist = TRUE,
    nCores = 10
)

thresSPE <- mean(n$thres)
bndwSPE <- mean(n$bndw)

# calculate the observation window based on the density
allStructs <- reconstructShapeDensitySPE(
    spe,
    marks = "type",
    imageCol = "sample_id",
    markSelect = NULL,
    bndw = bndwSPE,
    thres = thresSPE,
    nCores = 10
)
```

## ppp object from the window estimated with the intensity of the points
```{r}
combined_poly <- sf::st_union(allStructs)

# 2. Convert to owin object
window_estint <- as.owin(combined_poly)
ppp_int <- ppp(x = as.vector(coords[,"x"]), y = as.vector(coords[,"y"]), window = window_estint, marks = spe$type)
```

Almost 10k points were rejected as outside the window

## ppp object from the window estimated as a convex hull
```{r}
ppp_obj <- ppp(x = as.vector(coords[,"x"]), y = as.vector(coords[,"y"]), window = window_polygon, marks = spe$type)
```

No problem here with points outside the window

### grid creation for both the windows

Let's set a grid size so we can have smaller tiles and more local informations. Fixing the size to 2000 pixel and with mpp of 0.25 (i guess) we have tiles of 500 micron 

```{r}
tiles_creation <- function(window, tile_size = 2000){
  
  x_range <- diff(window$xrange)
  y_range <- diff(window$yrange)
  
  nx_calc <- ceiling(x_range / tile_size)
  ny_calc <- ceiling(y_range / tile_size)
  
  grid_rect <- quadrats(window, nx = nx_calc, ny = ny_calc)
  
  # 4. Clip to the polygon (Same logic as before)
  grid_tiles <- tiles(grid_rect)
  clipped_tiles <- lapply(grid_tiles, function(w) intersect.owin(w, window))
  
  # 5. Filter empty tiles and recreate tessellation
  valid_tiles <- clipped_tiles[sapply(clipped_tiles, function(x) !is.empty(x))]
  final_tess <- tess(tiles = valid_tiles)
  final_tess
}
```

Intensity estimated

```{r}
final_tess <- tiles_creation(window_estint, tile_size = 2000)

split_points <- split(ppp_int, f = final_tess)
# 5. Verify the shape
# plot(final_tess, main = "Grid clipped to Polygon Shape")
```

Decide a threshold to remove (almost) empty tiles
```{r}
npoints_tess <- map_dbl(split_points, npoints)
length(npoints_tess)
summary(npoints_tess)
```

Cut at the 1st Quartile of the npoints distribution
```{r}
ppp_intsub <- split_points[sapply(split_points, npoints) >= 428]
ppp_intsub
```

Do the same for the other window 
```{r}
grid_tess <- tiles_creation(window_polygon, tile_size = 2000)

# Cut the points into the quadrats
split_points <- split(ppp_obj, f = grid_tess)
```

Keep filtering at the first interquantile of before
```{r}
# Example: Check how many points are in the first FOV
npoints_tess <- map_dbl(split_points, npoints)
length(npoints_tess)
summary(npoints_tess)
# filter tess with npoints lower then 10k
ppp_sub <- split_points[sapply(split_points, npoints) >= 428]
ppp_sub
```

Now I have the 2 kind of object - I want to check if the marks are proportional and then keep only the most similar tiles in term of distribution of marks and quantity 
And save it 
```{r}
ppp_all <- list("ppp_estint" = ppp_intsub, "ppp_win" = ppp_sub)
# 
saveRDS(ppp_all, file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/ppp_list_of_both_ppp.RDS")

ppp_all <- readRDS("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/ppp_list_of_both_ppp.RDS")
```


# Esploratory graphics
```{r}
# 1. Create the data frame iterating through the nested lists
df_counts <- imap_dfr(ppp_all, function(ppp_list, source_name) {
  
  # Iterate through the tiles in the current list
  imap_dfr(ppp_list, function(ppp_obj, tile_name) {
    
    # Extract marks and count them
    marks_vec <- marks(ppp_obj)
    
    # Handle cases where a tile might be empty
    if (npoints(ppp_obj) == 0) {
      return(tibble(Mark = character(), Count = integer(), Tile = character()))
    }
    
    # Create the summary tibble
    table(marks_vec) |> 
      as_tibble() |> 
      rename(Mark = marks_vec, Count = n) |> 
      mutate(Tile = tile_name)
      
  }) |> 
    mutate(Source = source_name) # Add the top-level list name (ppp_estint, ppp_win)
})

# 2. Plot the results
ggplot(df_counts, aes(x = Tile, y = Count, fill = Mark)) +
  geom_col(position = "dodge") +
  facet_wrap(~Source, ncol = 1, scales = "free_y") + # Separate plots for each source
  theme_minimal() +
  labs(
    title = "Point Counts per Mark by Tile and Source",
    x = "Tile ID",
    y = "Number of Points"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# 3. Plot proportions using position = "fill"
ggplot(df_counts, aes(x = Tile, y = Count, fill = Mark)) +
  geom_bar(stat = "identity", position = "fill") + 
  scale_y_continuous(labels = scales::percent) + # Converts y-axis to %
  facet_wrap(~Source, ncol = 1, scales = "free_y") + # Separate plots for each source
  theme_minimal() +
  labs(title = "Proportion of Marks per Tile",
       x = "Tile",
       y = "Proportion",
       fill = "Mark Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# 2. Iterate through the top-level list (estint vs win)
# iwalk(ppp_all, function(ppp_list, source_name) {
# 
#   # Create a specific sub-folder for this source (e.g., plots/ppp_estint)
#   save_dir <- file.path("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/figures/spatial_plots", source_name)
#   dir.create(save_dir, showWarnings = FALSE)
#   
#   # 3. Iterate through the tiles within this source
#   iwalk(ppp_list, function(current_ppp, tile_name) {
#     
#     # Extract data
#     df_points <- as.data.frame(current_ppp)
#     
#     # Generate the plot
#     p <- ggplot(df_points, aes(x = x, y = y, color = marks)) +
#       geom_point(size = 0.05, alpha = 0.6) + 
#       coord_fixed() + 
#       theme_minimal() +
#       labs(title = paste(source_name, "-", tile_name)) + # Add source to title
#       theme(
#         axis.text = element_blank(),
#         axis.ticks = element_blank(),
#         axis.title = element_blank(),
#         panel.grid = element_blank()
#       ) + 
#       guides(color = guide_legend(override.aes = list(size = 5, alpha = 1)))
#     
#     # 4. Construct filename and Save
#     # Sanitizing filename to remove potential illegal characters
#     safe_tile_name <- gsub("[^[:alnum:]]", "_", tile_name) 
#     file_path <- file.path(save_dir, paste0(safe_tile_name, ".png"))
#     
#     ggsave(filename = file_path, plot = p, width = 12, height = 12, dpi = 400)
#     
#     message(paste("Saved:", file_path))
#   })
# })
```

Now i need to filter the tiles and keep the most similar in term of quantity of marks
I'm interested in the neoplastic and stromal for now.
```{r}
tile_ratios <- df_counts |>
  # Keep only the relevant marks
  filter(Mark %in% c("neoplastic", "stromal")) |>
  
  # Pivot to have columns: neoplastic, stromal
  tidyr::pivot_wider(names_from = Mark, values_from = Count, values_fill = 0) |>
  
  # Calculate the proportion of neoplastic cells (0 to 1)
  # This normalizes the data so 100k:1k is treated the same as 100:1
  mutate(
    neo_stroma_prop = neoplastic / stromal,
  ) #|>
  
  # Filter out empty or very sparse tiles to avoid division by zero errors
  #filter(Total > 10) 

# 2. Determine the "Target" Proportion
# You can pick the Median of your dataset, or a specific number (e.g., 0.90)
target_prop <- median(tile_ratios$neo_stroma_prop, na.rm = TRUE)
tolerance <- 0.2 # How much deviation is allowed? (e.g., +/- 5%)

cat("Target neoplastic stromal Proportion:", round(target_prop, 3), "\n")
cat("Range:", round(target_prop - tolerance, 3), "to", round(target_prop + tolerance, 3), "\n")

# 3. Filter the Tiles
selected_tiles <- tile_ratios |>
  filter(
    neo_stroma_prop >= (target_prop * (1 - tolerance)),
    neo_stroma_prop <= (target_prop * (1 + tolerance))
  )

# 4. View the selected tiles
print(selected_tiles)

# --- OPTIONAL: Visualize the Distribution of Ratios ---
# This helps you decide if your tolerance is too strict or too loose
ggplot(tile_ratios, aes(x = neo_stroma_prop)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = target_prop, color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = c((target_prop * (1 - tolerance)), (target_prop * (1 + tolerance))), color = "red", linetype = "dotted", linewidth = 0.8) +
  annotate("rect", xmin = (target_prop * (1 - tolerance)), xmax = (target_prop * (1 + tolerance)), 
           ymin = -Inf, ymax = Inf, fill = "red", alpha = 0.1) +
  theme_minimal() +
  labs(title = "Distribution of Neoplastic/Stromal Ratios",
       subtitle = paste0("Target (Median): ", round(target_prop, 2), 
                         " | Accepted Range: ", round((target_prop * (1 - tolerance)), 2), "-", round((target_prop * (1 + tolerance)), 2)),
       x = "Ratio (Neoplastic / Stromal)",
       y = "Count of Tiles")
```

Subset to keep only the 4 tiles for the 2 windows type
```{r}
ppp_all_filt <- imap(ppp_all, \(current_ppp_list, source_name){
  # names of the tiles to keep for each source
  names_4each_source <- selected_tiles |> 
    filter(Source == source_name) |> 
    pull(Tile)
  # names of the tiles to filter to
  tiles_to_keep <- intersect(names(current_ppp_list), names_4each_source)
  return(current_ppp_list[tiles_to_keep])
})

ppp_all_filt

# 3. Check the results
# See how many tiles remain in each list
map(ppp_all_filt, length)
```


# Homogeneus ppp

L function between neoplastic and stromal
```{r}
Lcross_homo <- imap(ppp_all_filt, \(current_ppp_list, source_name) {
  
  imap(current_ppp_list, \(ppp_obj, tile_name) {
     # Calculate Lcross
     L <- Lcross(ppp_obj, i = "neoplastic", j = "stromal", correction = "best")
     L 
  }) 
})
```

and pair correlation function

```{r}
pcfcross_homo <- imap(Lcross_homo, \(current_L_list, source_name) {
  
  imap(current_L_list, \(Lcross, tile_name) {
     # Calculate Lcross
     pcf <- pcf(Lcross, method = "b")
     pcf
  }) 
})
```

## Plot of the curvers

```{r}
df_Lcross_homo <- imap_dfr(Lcross_homo, function(tile_list, source_name) {
  
  imap_dfr(tile_list, function(L_obj, tile_name) {
    
    # Convert fv object to dataframe
    as.data.frame(L_obj) |> 
      select(r, theo, matches("iso|trans|border")) |>  # Select r, theo, and the edge correction used
      rename_with(~"observed", matches("iso|trans|border")) |>  # Rename correction col to 'observed'
      mutate(
        Source = source_name,
        Tile = tile_name,
        # Calculate L(r) - r (Besag's transformation) for easier interpretation
        # If L(r) - r > 0 => Clustering (Attraction)
        # If L(r) - r < 0 => Dispersion (Repulsion)
        L_centered = observed - r 
      )
  })
})
```


```{r}
ggplot(df_Lcross_homo, aes(x = r, y = L_centered, group = Tile)) +
  # Add the theoretical Poisson line (y=0 because we did L(r)-r)
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  
  # Plot all individual tile curves (faintly)
  geom_line(alpha = 0.3, color = "steelblue") +
  
  # Split by Source
  facet_wrap(~Source) +
  
  theme_minimal() +
  labs(
    title = "L-cross Function (Besag's L(r) - r)",
    subtitle = "Positive = Association + , Negative = Association -",
    x = "Distance (r)",
    y = "L(r) - r"
  )
```

```{r}
df_pcfcross_homo <- imap_dfr(pcfcross_homo, function(tile_list, source_name) {
  
  imap_dfr(tile_list, function(pcf_obj, tile_name) {
    
    # Convert fv object to dataframe
    as.data.frame(pcf_obj) |> 
      select(r, theo, pcf) |>  # Select r, theo, and the edge correction used
      mutate(
        Source = source_name,
        Tile = tile_name,
        # Calculate L(r) - r (Besag's transformation) for easier interpretation
        # If L(r) - r > 0 => Clustering (Attraction)
        # If L(r) - r < 0 => Dispersion (Repulsion)
        pcf = pcf
      )
  })
})
```

```{r}
ggplot(df_pcfcross_homo, aes(x = r, y = pcf, group = Tile)) +
  # Add the theoretical Poisson line (y=0 because we did L(r)-r)
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  
  # Plot all individual tile curves (faintly)
  geom_line(alpha = 0.3, color = "steelblue") +
  
  # Split by Source
  facet_wrap(~Source) +
  
  theme_minimal() +
  labs(
    title = "L-cross Function (Besag's L(r) - r)",
    subtitle = "Positive = Association + , Negative = Association -",
    x = "Distance (r)",
    y = "L(r) - r"
  )
```

```{r}
library(ggrepel)
label_data <- df_Lcross_homo %>%
  group_by(Source, Tile) %>%
  filter(r == max(r)) %>% # Select the end of the curve
  ungroup()

# 2. Plot with Colors and Labels
ggplot(df_Lcross_homo, aes(x = r, y = L_centered, group = Tile, color = Tile)) +
  
  # Theoretical Line
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
  
  # Curves (Colored by Tile)
  geom_line(alpha = 0.8, linewidth = 0.7) +
  
  # Labels at the end of the curves
  geom_text_repel(
    data = label_data,
    aes(label = Tile),
    size = 3,
    direction = "y",      # Try to stack them vertically if they overlap
    segment.size = 0.2,   # Thin lines connecting label to curve
    nudge_x = 0.1,        # Push labels slightly to the right
    max.overlaps = 20     # Allow many labels
  ) +
  
  # Split by Source
  facet_wrap(~Source) +
  
  theme_minimal() +
  labs(
    title = "L-cross Function (Besag's L(r) - r)",
    subtitle = "Positive = Attraction, Negative = Repulsion",
    x = "Distance (r)",
    y = "L(r) - r"
  ) +
  # Hide the legend because the direct labels make it redundant
  theme(legend.position = "none") 
```


# Intensity estimate for the tiles

I want to estimate intensity based on the 2 marks I'll study, so neoplastic and stromal. And then use it as a proxy to model Ripley K function of an inhomogeneus process with intensity estimate on the 2 marks kernel. I can try to estimate the kernel intensity with different parameters and see what does it change. also maybe Hypotesize different kinds of second-order organization?





