---
title: "Pair Correlation function for TCGA data"
author: "Valerio Reffo"
date: "2026-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(anndataR)
library(ggspavis)
library(patchwork)
library(spatstat.geom)
library(spatstat)
library(sosta)
library(purrr)
library(readxl)
library(dplyr)
library(spatialFDA)
library(fda)
```

Function to estimate Lcross function for homogeneus point pattern between 2 marks and with downsample
```{r}
pcfcross_personal <- function(ppp_obj, mark_i, mark_j, downsample = TRUE, n_down = 10000, ...) {
  
  # 1. Get the indices of the specific cell types dynamically
  idx_i <- which(marks(ppp_obj) == mark_i)
  idx_j <- which(marks(ppp_obj) == mark_j)
  
  # Safety check: ensure both cell types actually exist in the image!
  if(length(idx_i) < 2 | length(idx_j) < 2) {
    warning(paste("Not enough points for", mark_i, "and/or", mark_j, "to compute Lcross. Returning NULL."))
    return(NULL)
  }

  # 2. Handle Downsampling
  if (downsample) {
    # Randomly sample a maximum of n_down cells of each type
    keep_i <- sample(idx_i, min(length(idx_i), n_down))
    keep_j <- sample(idx_j, min(length(idx_j), n_down))
    
    # Subset the ppp object
    ppp_sub <- ppp_obj[c(keep_i, keep_j)]
    
  } else {
    # Even if we don't downsample, subsetting to JUST the two marks of interest 
    # saves a massive amount of memory for spatstat calculation
    ppp_sub <- ppp_obj[c(idx_i, idx_j)]
  }

  # 3. Estimate of bw with bw.diggle (likelihood estimate)
  sigma_val <- bw.diggle(ppp_sub)
  # 4. Calculate pcfcross dynamically using the provided mark names, 
  #    preferred divisor = "d" 
  pcf <- pcfcross(ppp_sub, 
              i = mark_i, 
              j = mark_j, 
              bw = sigma_val, ...)
  
  return(pcf)
}
```


```{r}

r_values <- seq(0, 100, length.out = 40)

pcfcross_benign <- map(ppp_list, \(ppp_obj) {
  pcfcross_personal(ppp_obj, mark_i = "neoplastic", mark_j = "benign epithelial", 
                  downsample = TRUE, n_down = 50000, correction = "translate",
                  kernel = "epanechnikov", divisor = "d", rmax = 100)
})

Lcross_stroma <- map(ppp_list, \(ppp_obj) {
  Lcross_personal(ppp_obj, mark_i = "neoplastic", mark_j = "stromal", 
                  downsample = TRUE, n_down = 50000, rmax = 100, correction = "border", r = r_values)
})
```

```{r}
walk(pcfcross_benign, \(pcf) {plot(pcf, xlim = c(0, 1000))})
```



Plot the results
```{r}
plot_fv_list <- function(list_of_funcs, correction = "border") {
  
  # 1. Run the map_dfr block dynamically on whatever list you provide
  df_lcross <- map_dfr(names(list_of_funcs), function(slide_name) {
    
    # Extract the specific L-function object from the provided list
    L_obj <- list_of_funcs[[slide_name]]
    
    as.data.frame(L_obj) |> 
      mutate(
        Slide = slide_name,
        
        # Use .data[[correction]] to dynamically grab the 'border' column (or 'iso', etc.)
        L_r = .data[[correction]],  
        
        # Extract the marks names exactly as you designed
        mark_i = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][1]),
        mark_j = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][2])
      )
  }, .id = "sample_ID")
  
  # 2. Build the ggplot
  p <- ggplot(df_lcross, aes(x = r, y = L_r, group = Slide, color = Slide)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(alpha = 0.7, linewidth = 0.6) + # Using linewidth instead of size for modern ggplot2
    theme_minimal() +
    labs(
      title = paste0("pcf-cross Function: ", df_lcross$mark_i[1], " vs ", df_lcross$mark_j[1]),
      subtitle = "Centered L-function (L(r) - r). Positive = Attraction, Negative = Repulsion",
      x = "Radius (r)",
      y = "pcf(r)",
      color = "Slide"
    ) +
    # Truncate the legend strictly to the first 12 characters (TCGA-XX-XXXX)
    scale_color_discrete(labels = function(x) substr(x, 1, 12)) +
    theme(legend.position = "right")
  
  return(p)
}
```


Plot of the lcross functions for neoplastic vs benign epitelial and stromal
```{r}
(plcross_nb <- plot_fv_list(Lcross_benign, correction = "trans"))
(plcross_ns <- plot_fv_list(Lcross_stroma))
```

# Functional Data Analysis


# generic function for smoothing funs
```{r}
fsmooth <- function(fcross_list, fun, M = 6, genLfun.fd = 4, centrata = F, ...)
{

  # 1. Calculate raw L-functions (Common for both)
  r <- fcross_list[[1]]$r
  mat_fun <- sapply(fcross_list, function(x) x$border)
  
  # 2. Apply centering ONLY if requested
  if(centrata == TRUE){
    mat_fun <- sweep(mat_fun, MARGIN = 1, STATS = r, FUN = "-")
  }

  # 3. Create Basis and Smooth
  K <- length(r) + M - 2
  genbasis <- create.bspline.basis(range(r), K, M, r)
  genfdPar <- fdPar(genbasis, genLfun.fd, lambda = 1e-11)
  
  genfdSmooth <- smooth.basis(r, mat_fun, genfdPar)
  Genfun.fd <- genfdSmooth$fd
  
  # 4. Set Names
  fdn_dist <- paste("Distanza (da 0 a", max(r), "micron)")
  fdnames <- list(fdn_dist,
                  "Gene" = colnames(mat_fun), # Crucial: use names of surviving columns
                  "Funzione")
  Genfun.fd$fdnames <- fdnames
  
  return(list(fd = Genfun.fd, fdPar = genfdPar, 
              basis = genbasis, mat = mat_fun, r = r))
}
```



```{r}
Lfun_benign <- fsmooth(Lcross_benign, centrata = T)
Lfun_stroma <- fsmooth(Lcross_stroma, centrata = T)
```

```{r}
plot(Lfun_benign$fd)
lines(c(0,350), c(0,0), col = "gold", lwd = 2)

plot(Lfun_stroma$fd)
lines(c(0,350), c(0,0), col = "gold", lwd = 2)
```

## Comparison between neoplastic and benign.epithelial

### Functional PCA

Calcoliamo fpca per 4 componenti e mostriamo il risultato sulla rotazione VARIMAX

```{r}
fpca <- pca.fd(Lfun_benign$fd, nharm = 4, Lfun_benign$fdPar)
rownames(fpca$scores) <- names(Lcross_benign) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca)
```

### fPCA plot ruotate per VARIMAX

The plots below show the mean function and the effects of adding (+) and subtracting (−) a
suitable multiple of the first (resp. second) principal component function.

```{r}
plot.pca.fd(fpcaVarmx, cex.main=0.9)
```

## Clustering

Applichiamo due metodi di clustering sugli scores delle prime 4 componenti.
Clustering gerarchico con metodo Ward e kmeans

```{r}
set.seed(123)
fpca_km <- kmeans(fpca$scores[,c(1,4)], 4)
fpca_hc <- hclust(dist(fpca$scores[,c(1,4)]), method = "ward.D2")
memb <- cutree(fpca_hc, k = 4)
```

Grafico del dendogramma per il clustering gerarchico, scegliamo di dividere in 5 cluster.

```{r}
dhc <- as.dendrogram(fpca_hc)
plot(dhc, cex = 0.1, horiz = T)
#ggdendrogram(dhc, rotate = TRUE, size = 2)
table(kmean = fpca_km$cluster)
table(ward.d2 = memb)
```

```{r}
plot(fpca$scores[,1], fpca$scores[,2], col = fpca_km$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca$scores[,1], fpca$scores[,2], col = memb, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

```{r}
df_A <- data.frame(
  patientID = c(
    "TCGA-OY-A56P", "TCGA-23-1021", "TCGA-23-2072", "TCGA-23-1031", "TCGA-23-1117", # Cluster 1
    "TCGA-25-2401", "TCGA-25-2399", "TCGA-25-1625", "TCGA-25-1328", "TCGA-25-1871", # Cluster 2
    "TCGA-25-1315", "TCGA-25-1326", "TCGA-25-2391", "TCGA-25-1635", "TCGA-25-1627", # Cluster 3
    "TCGA-57-1582", "TCGA-13-A5FU", "TCGA-57-1993", "TCGA-57-1994", "TCGA-57-1583"  # Cluster 4
  ),
  cluster_ilaria = c(
    rep(1, 5), 
    rep(2, 5), 
    rep(3, 5), 
    rep(4, 5)
  )
)

# 3. Clean Dataset B to match IDs
# We extract the first 12 characters (e.g., "TCGA-25-1326")
df_clust <- data.frame(
  patientID = substr(names(vec_kmean), 1, 12),
  cluster_kmean = fpca_km$cluster,
  cluster_ward = memb
) |> print()

# 4. Merge to find correspondence
comparison_benign <- inner_join(df_A, df_clust, by = "patientID") |> 
  arrange(cluster_kmean, patientID)

# 5. Display the Result
print(comparison_benign)

table("ilaria_clust" = comparison_benign$cluster_ilaria, "valerio_clust" = comparison_benign$cluster_kmean)
```

## Comparison between neoplastic and stromal 

### Functional PCA

Calcoliamo fpca per 4 componenti e mostriamo il risultato sulla rotazione VARIMAX

```{r}
fpca <- pca.fd(Lfun_stroma$fd, nharm = 4, Lfun_stroma$fdPar)
rownames(fpca$scores) <- names(Lcross_stroma) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca)
```

### fPCA plot ruotate per VARIMAX

The plots below show the mean function and the effects of adding (+) and subtracting (−) a
suitable multiple of the first (resp. second) principal component function.

```{r}
plot.pca.fd(fpcaVarmx, cex.main=0.9)
```

## Clustering

Applichiamo due metodi di clustering sugli scores delle prime 4 componenti.
Clustering gerarchico con metodo Ward e kmeans

```{r}
set.seed(123)
fpca_km <- kmeans(fpca$scores[,c(1,4)], 4)
fpca_hc <- hclust(dist(fpca$scores[,c(1,4)]), method = "ward.D2")
memb <- cutree(fpca_hc, k = 4)
```

Grafico del dendogramma per il clustering gerarchico, scegliamo di dividere in 5 cluster.

```{r}
dhc <- as.dendrogram(fpca_hc)
plot(dhc, cex = 0.1, horiz = T)
#ggdendrogram(dhc, rotate = TRUE, size = 2)
table(kmean = fpca_km$cluster)
table(ward.d2 = memb)
```