---
title: "Embedding PCA"
author: "Valerio Reffo"
date: "2026-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(anndataR)
library(ggspavis)
library(patchwork)
library(spatstat.geom)
library(spatstat)
library(sosta)
library(data.table)
library(purrr)
#library(ImageFeatureTCGA)
library(Rtsne)
library(ggplot2)
#source("~/repositories/spatialPointProcess_clustering/helper_function.R", echo = TRUE)
# survival analysis
library(survival)
library(survminer)
```

```{r}
pgp_emb <- fread("/mnt/ganimede/shared/imageFeatureTCGA_OV/provgigapath/slide_level/TCGA-WR-A838-01Z-00-DX1.5FE22DE4-CEFB-45F6-9299-505023A8F3BA.csv.gz", 
                  quote = "\"", 
                  fill = TRUE)
# pgp_emb2 <- read.csv("/mnt/ganimede/shared/imageFeatureTCGA_OV/provgigapath/slide_level/TCGA-WR-A838-01Z-00-DX1.5FE22DE4-CEFB-45F6-9299-505023A8F3BA.csv.gz")
```

# Codice di copilot per processare gli embeddings e fare PCA

1. Define folder and get all csv.gz files
```{r}
folder_path <- "/mnt/ganimede/shared/imageFeatureTCGA_OV/provgigapath/slide_level"
file_list <- list.files(folder_path, pattern = "\\.csv\\.gz$", full.names = TRUE)
```


2. Function to clean the PyTorch string and convert to numeric vector
```{r}
parse_tensor_string <- function(tensor_str) {
  # Remove "tensor([[", "]])", and newlines (\n)
  clean_str <- gsub("tensor\\(\\[\\[|\\]\\]\\)", "", tensor_str)
  clean_str <- gsub("\n", "", clean_str)
  
  # Split by comma
  split_str <- strsplit(clean_str, ",")[[1]]
  
  # Convert to numeric (R automatically handles scientific notation like 5.4e-02)
  num_vec <- as.numeric(trimws(split_str))
  
  return(num_vec)
}
```


3. Function to process a single file
```{r}
process_slide <- function(file_path) {
  # Read the single row of data
  dt <- fread(file_path, quote = "\"", fill = TRUE, showProgress = FALSE)
  
  slide_id <- dt$slide_name[1]
  
  # --- Extract just the last layer (Recommended) ---
  features <- parse_tensor_string(dt$last_layer_embed[1])
  
  # --- ALTERNATIVE: If you truly want ALL layers concatenated (Layers 0 to 12) ---
  # embed_cols <- c(paste0("layer_", 0:12, "_embed"), "last_layer_embed")
  # parsed_layers <- lapply(dt[, ..embed_cols], parse_tensor_string)
  # features <- unlist(parsed_layers)
  
  return(list(slide_name = slide_id, features = features))
}
```


4. Loop over all files
```{r}
# (Tip: test this on the first 5 files first by using file_list[1:5])
cat("Processing", length(file_list), "slides...\n")
parsed_data <- map(file_list, process_slide)

# 5. Build the final matrix
# Bind all feature vectors as rows
pca_matrix <- do.call(rbind, map(parsed_data, function(x) x$features))

# Set the row names to the slide names
rownames(pca_matrix) <- sapply(parsed_data, function(x) x$slide_name)

cat("Done! Matrix dimensions:", dim(pca_matrix)[1], "rows (slides) x", dim(pca_matrix)[2], "columns (features)\n")
```

```{r}
# Run PCA (center and scale the features)
pca_res <- prcomp(pca_matrix, center = T, scale. = T)

# Plot the variance explained
plot(pca_res, type = "l", main = "PCA Variance")

# Get your PC scores for plotting (e.g., PC1 and PC2)
pc_scores <- as.data.frame(pca_res$x)
rownames(pc_scores) <- rownames(pca_res$x)
head(pc_scores[, 1:2])
```

```{r}
var_explained <- summary(pca_res)$importance[2, 1:2] * 100
var_explained <- round(var_explained, 1) # Round to 1 decimal place
plot(
  x = pca_res$x[, "PC1"], 
  y = pca_res$x[, "PC2"],
  xlab = paste0("PC1 (", var_explained[1], "%)"),
  ylab = paste0("PC2 (", var_explained[2], "%)"),
  main = "PCA of Prov-GigaPath Slide Embeddings",
  pch = 19,                     # Solid circles
  col = rgb(0.1, 0.4, 0.8, 0.6),# Nice translucent blue
  cex = 1.2                     # Slightly larger points
)

# 3. Add a grid in the background to make it easier to read
grid(col = "lightgray", lty = "dotted")
```

# t-SNE

```{r}
set.seed(42) # t-SNE is stochastic, set seed for reproducibility
tsne_out <- Rtsne(pc_scores[,1:30], dims = 2, perplexity = 30, verbose = TRUE, check_duplicates = FALSE)

# 3. Prepare data for plotting
# Combine the t-SNE coordinates with your original metadata (e.g., labels)
tsne_plot_data <- data.frame(
  TSNE1 = tsne_out$Y[, 1],
  TSNE2 = tsne_out$Y[, 2]
  # Assuming 'my_labels' is a vector of your groups/classes corresponding to the rows
)

# 4. Plot with ggplot2
ggplot(tsne_plot_data, aes(x = TSNE1, y = TSNE2)) +
  geom_point(size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "t-SNE Projection (from PCA)",
       subtitle = "Perplexity = 30",
       x = "t-SNE Dimension 1",
       y = "t-SNE Dimension 2") +
  scale_color_viridis_d() # Optional: nice color palette
```

# K-means on the embeddings

```{r}
set.seed(123)
clusters <- kmeans(pca_matrix, 4)
table(clusters$cluster)
```

plot of the clusters on the tSNE
```{r}
tsne_plot_data <- data.frame(
  TSNE1 = tsne_out$Y[, 1],
  TSNE2 = tsne_out$Y[, 2],
  clusters = as.factor(clusters$cluster)
  # Assuming 'my_labels' is a vector of your groups/classes corresponding to the rows
)

# 4. Plot with ggplot2
ggplot(tsne_plot_data, aes(x = TSNE1, y = TSNE2, color = clusters)) +
  geom_point(size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "t-SNE Projection (from PCA)",
       subtitle = "Perplexity = 30",
       x = "t-SNE Dimension 1",
       y = "t-SNE Dimension 2")
 
```

PCA plot colored by clusters
```{r}
pca_data <- as.data.frame(pca_res$x)

# Ensure the row names of pca_data match the names in clusters$cluster if they aren't in the exact same order.
# If they are already aligned, you can just assign: pca_data$Cluster <- as.factor(clusters$cluster)
# However, merging by name is safer:
pca_data$SampleID <- rownames(pca_data)

# Convert clusters to a dataframe for merging
cluster_df <- data.frame(
  SampleID = names(clusters$cluster),
  Cluster = as.factor(clusters$cluster) # Convert to factor for categorical coloring
)

# Merge to ensure correct mapping
plot_data <- left_join(pca_data, cluster_df, by = "SampleID")

# 2. Get Variance Explained (reusing your logic)
var_explained <- summary(pca_res)$importance[2, 1:2] * 100
var_explained <- round(var_explained, 1)

# 3. Create the Plot
ggplot(plot_data, aes(x = PC1, y = PC2, color = Cluster)) +
  # Add points with some transparency
  geom_point(size = 3, alpha = 0.7) + 
  
  # Optional: Add ellipses to visualize cluster boundaries
  #stat_ellipse(level = 0.95, linetype = "dashed", size = 0.5, show.legend = FALSE) +
  
  # Labels and Titles
  labs(
    title = "PCA of Prov-GigaPath Slide Embeddings",
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)"),
    color = "Cluster"
  ) +
  
  # Theme and Colors
  theme_minimal() +
  scale_color_viridis_d(option = "D") + # 'viridis' is colorblind-friendly and distinct
  theme(
    panel.grid.major = element_line(color = "lightgray", linetype = "dotted"),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```
```{r}
head(plot_data)
which.min(plot_data$PC1)
which.max(plot_data$PC1)
which.max(plot_data$PC2)
which.min(plot_data$PC2)

```


# Survival analysis

## Extract survival times

file with info about patients
```{r}
library(ExperimentHub)
library(curatedTCGAData)
ov_meta <- curatedTCGAData(diseaseCode = "OV", assays = "*", version = "1.1.38", dry.run = FALSE)

ov_meta |> head()

ov_meta <- colData(ov_meta) %>% as.data.frame

# select clinical meta data
columns_to_select <- c(
  "patientID",                             # patient id
  "vital_status",                          # Survival status
  "years_to_birth",                          # Birth date
  "days_to_death",                         # Time variable for deceased patients
  "days_to_last_followup",                 # Time variable for censored patients
  "patient.age_at_initial_pathologic_diagnosis", # Age 
  "gender",                                # Gender/Sex
  "ethnicity",                             # Ethnicity
  "patient.anatomic_neoplasm_subdivision" # Anatomic site 
  # TBD adding other vars of interest

)

# Select the vars
ov_meta_select <- ov_meta[, columns_to_select, drop = FALSE]
ov_meta_select |> head()
```


samples names, there are 107 samples for 106 patient
```{r}

file_list <- list.files("/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/h5ad", pattern = "\\.h5ad$", full.names = TRUE)
(sample_names <- tools::file_path_sans_ext(basename(file_list)))
(patient_ids_from_files <- substr(sample_names, 1, 12))
```

```{r}
ov_meta_matched <- ov_meta_select |> 
  filter(patientID %in% patient_ids_from_files)

# Check the result
head(ov_meta_matched)
```

cluster setup
```{r}
slide_names <- rownames(clusters) 
if (is.null(slide_names)) slide_names <- names(clusters$cluster)

# 2. Create a clean dataframe with Slide Name, Patient ID, and Cluster
ov_df <- data.frame(
  slide_name = slide_names,
  # Extract the first 12 characters to get the patientID (e.g., "TCGA-13-A5FT")
  patientID = substr(slide_names, 1, 12), 
  cluster = clusters$cluster,
  stringsAsFactors = FALSE
)

# 3. Merge the cluster data with the metadata
# This will automatically duplicate the metadata rows for patients with multiple slides!
ov_meta_matched <- ov_df |> 
  left_join(ov_meta_select, by = "patientID")

# 4. Optional: Set the row names of the new dataframe to the slide names
rownames(ov_meta_matched) <- ov_meta_matched$slide_name

# Check the result!
head(ov_meta_matched)
```


## Survival analysis 

Variables setup 
```{r}
ov_meta_matched <- ov_meta_matched |> 
  mutate(
    time = ifelse(!is.na(days_to_death), days_to_death, days_to_last_followup),
    event = ifelse(!is.na(days_to_death), 1, 0)
  )
```

```{r}
fit <- survfit(Surv(time, vital_status) ~ cluster, data = ov_meta_matched)
km_plot <- ggsurvplot(fit, data = ov_meta_matched, risk.table = TRUE, pval = TRUE, xlim = c(0,3000))
pairwise_survdiff(Surv(time, vital_status) ~ cluster, data = ov_meta_matched)
```

## check the groups

Checking if I made some mistakes in copying clusters elements
```{r}
cluster_patients_df <- ov_meta_matched |> 
  group_by(cluster) |> 
  summarise(
    # paste(unique(...)) ensures that if a patient has 2 slides in the same cluster, 
    # their ID only prints once in the list
    patient_IDs = paste(unique(patientID), collapse = ", ")
  ) |> 
  arrange(cluster)

km_result_pats <- clusters

# Print the resulting dataframe
print(cluster_patients_df)

orig_slide_names <- names(clusters$cluster)
orig_clusters <- as.numeric(clusters$cluster)

# 2. Extract the patient IDs (first 12 characters) from those original slide names
orig_patient_ids <- substr(orig_slide_names, 1, 12)

# 3. Create a temporary dataframe from the original data, group, and collapse
orig_summary_df <- data.frame(
  cluster = orig_clusters,
  patientID = orig_patient_ids,
  stringsAsFactors = FALSE
) %>%
  group_by(cluster) %>%
  summarise(
    original_patient_IDs = paste(unique(patientID), collapse = ", ")
  )

# 4. Join this directly into your existing cluster_patients_df
cluster_patients_df <- cluster_patients_df %>%
  left_join(orig_summary_df, by = "cluster") %>%
  # Add a check to confirm they are exactly identical!
  mutate(is_match = (patient_IDs == original_patient_IDs))

# Print it to check!
print(cluster_patients_df)
```

# Saving list of patientID per clusters

```{r}
saveRDS(ov_meta_matched, file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/survival_dataset_&_metadata.rds")
```

```{r}
# Set a seed so you get the exact same random samples if you rerun the code
set.seed(42) 

random_5_per_cluster <- ov_meta_matched |>
  # 1. Keep only unique patients per cluster so we don't sample the same person twice
  distinct(cluster, patientID) |>
  # 2. Group by cluster
  group_by(cluster) |>
  # 3. Randomly pull exactly 5 rows (if a cluster has fewer than 5, it pulls all of them)
  slice_sample(n = 5) |>
  ungroup() |>
  # Keep just the columns you care about
  select(cluster, patientID)

# Print the final dataframe
print(random_5_per_cluster)

# If you need them as a simple flat vector of IDs to use later:
vector_of_ids <- random_5_per_cluster$patientID
```

```{r}
repeated_patients_check <- ov_meta_matched |>
  group_by(patientID) |>
  # Keep only patients that appear more than once (have multiple slides)
  filter(n() > 1) |> 
  summarise(
    num_slides = n(),
    # Count how many unique clusters this patient's slides were assigned to
    num_unique_clusters = n_distinct(cluster), 
    # List the exact clusters assigned
    assigned_clusters = paste(cluster, collapse = ", "),
    # List the slides just in case you want to check them
    slide_names = paste(slide_name, collapse = ", ")
  ) |>
  # If num_unique_clusters == 1, then all their slides are in the same group!
  mutate(all_in_same_group = (num_unique_clusters == 1)) |>
  arrange(all_in_same_group) # Put any FALSE ones at the top so they are easy to spot

```
Il paziente ripetente è giustamente assegnato solo al cluster 4. 


# Filtering images with double tissues
```{r}
good_sample_names <- c(
  "TCGA-13-A5FU-01Z-00-DX1.9AD9E4B9-3F87-4879-BC0F-148B12C09036",
  "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D",
  "TCGA-23-1022-01Z-00-DX1.AF9E523E-CB0F-4AB5-AD43-C96731BF9141",
  "TCGA-23-1023-01Z-00-DX1.0C96E118-A4D9-4A9A-B95E-C0AA114D2483",
  "TCGA-23-1024-01Z-00-DX1.B9194D3F-C6F4-4FC8-B0CA-6E347FF4F885",
  "TCGA-23-1026-01Z-00-DX1.2875B4F7-D6B2-4C72-8A68-4E7C92D04BF0",
  "TCGA-23-1027-01Z-00-DX1.53F9DFF4-6811-4184-B2FD-1F6706B948FD",
  "TCGA-23-1028-01Z-00-DX1.117B4B7B-F796-4D33-A645-CD80E5C43E6D",
  "TCGA-23-1029-01Z-00-DX1.0044B39A-51B3-4F76-90A9-00CDF851DE2A",
  "TCGA-23-1030-01Z-00-DX1.FC93147F-FB55-4C62-B187-337A5C107096",
  "TCGA-23-1031-01Z-00-DX1.12B1996D-E5BB-424C-B59E-4BE7E1ED6BA1",
  "TCGA-23-1032-01Z-00-DX1.0C1074AE-D59C-4C74-B929-593AE89E5DC1",
  "TCGA-23-1107-01Z-00-DX1.5B9C1283-62B5-4329-873D-009FA939CA69",
  "TCGA-23-1110-01Z-00-DX1.61C2DBB8-7F5C-4878-BEDC-6AE061A2D7D4",
  "TCGA-23-1111-01Z-00-DX1.A99421E1-C73E-42AE-990F-E5E6D26EDDBE",
  "TCGA-23-1113-01Z-00-DX1.D72B1587-444A-4DEC-8D92-703699822BE2",
  "TCGA-23-1114-01Z-00-DX1.26CCA42E-4947-4318-A983-D3B31603482E",
  "TCGA-23-1116-01Z-00-DX1.FC5FC1E1-56DD-498C-AA9F-4434C6E733A8",
  "TCGA-23-1117-01Z-00-DX1.783914B6-B079-4828-B77B-4C6306FAE4C1",
  "TCGA-23-1118-01Z-00-DX1.D69E130E-33C7-4F8A-99AF-400B81378578",
  "TCGA-23-1119-01Z-00-DX1.44D3D70A-A421-4962-8385-2A3CF8E9880A",
  "TCGA-23-1120-01Z-00-DX1.59367B12-17F1-41AA-A6FB-8D940365176A",
  "TCGA-23-1121-01Z-00-DX1.E2F25441-32C3-46BF-A845-CB4FA787E8CB",
  "TCGA-23-1122-01Z-00-DX1.991C230C-E3B4-4DAC-9DCC-69E9FA7DF772",
  "TCGA-23-1123-01Z-00-DX1.764A8E5C-F2E7-441D-B9FB-23ECF7183740",
  "TCGA-23-1809-01Z-00-DX1.A8767DEC-EB1A-4EDC-906E-C0560A30455F",
  "TCGA-23-2072-01Z-00-DX1.478243FF-BFF0-48A4-ADEA-DE789331A50E",
  "TCGA-23-2077-01Z-00-DX1.9EB664FA-CC13-4A49-89BA-8A4573CE48F1",
  "TCGA-23-2078-01Z-00-DX1.E31A9C1E-00EF-4B0D-A15D-6A5C2D00E986",
  "TCGA-23-2081-01Z-00-DX1.4105A382-4C2B-4C62-AC93-CB21CC360254",
  "TCGA-23-2641-01Z-00-DX1.DB792236-39EC-4090-9938-B0915C3378A0",
  "TCGA-23-2643-01Z-00-DX1.E60DF03D-326E-4182-84DA-77263C4C52B3",
  "TCGA-23-2645-01Z-00-DX1.72C0AB89-9A64-4656-B03E-A5D65BA03188",
  "TCGA-23-2647-01Z-00-DX1.21E5D0D8-6BA8-4D49-BA70-EC228CF10387",
  "TCGA-23-2649-01Z-00-DX1.C94C3B32-D5F5-4A85-9F35-50B377FF9E75",
  "TCGA-25-1312-01Z-00-DX1.733EC7A7-0FC8-4DDC-B366-DF5A45D6BB4E",
  "TCGA-25-1313-01Z-00-DX1.5A2732D5-01D6-4285-9283-1839E1703C27",
  "TCGA-25-1314-01Z-00-DX1.E01C2237-E2D6-4D84-9FB4-624734F30375",
  "TCGA-25-1315-01Z-00-DX1.BC434CCE-7E21-479F-8A5D-1FC7F601C8C5",
  "TCGA-25-1316-01Z-00-DX1.A9B7F47C-5BDC-4B7D-BC9F-C4C9DBA663C8",
  "TCGA-25-1317-01Z-00-DX1.2E7A7AF9-4D04-4EBD-8C79-7117F76511D5",
  "TCGA-25-1318-01Z-00-DX1.742CB28E-DB0A-4BE1-8047-5265FE80F861",
  "TCGA-25-1319-01Z-00-DX1.71EFB946-ACAF-4BA6-8855-D336268D87F0",
  "TCGA-25-1320-01Z-00-DX1.30C154CE-C75B-4815-BF05-7D011B761993",
  "TCGA-25-1321-01Z-00-DX1.56C898D8-14CE-4C2D-B5C9-AC33FF91FE68",
  "TCGA-25-1322-01Z-00-DX1.20DA1932-908F-44E5-B2F2-43E41E06165B",
  "TCGA-25-1323-01Z-00-DX1.1B82D1C9-2310-4AAB-B531-45C1191E5FF8",
  "TCGA-25-1324-01Z-00-DX1.37D77658-1D25-49B2-8CDE-FC092EAA4CDF",
  "TCGA-25-1325-01Z-00-DX1.FA0DD977-D2F2-4319-B614-3EB20E966D97",
  "TCGA-25-1326-01Z-00-DX1.686E3548-8891-42ED-ABBA-BE3DA6D4CAAB",
  "TCGA-25-1328-01Z-00-DX1.A0145C96-19A9-4F2D-8570-ECF039F99405",
  "TCGA-25-1329-01Z-00-DX1.6DD994F3-E426-4749-8ECA-9A0B87FF04DC",
  "TCGA-25-1623-01Z-00-DX1.28D89405-5F27-47CF-ABC7-01AB494C28D2",
  "TCGA-25-1625-01Z-00-DX1.64A0B073-8EC8-4BE1-BECA-B956A6A96FCC",
  "TCGA-25-1626-01Z-00-DX1.2C6EA1FE-8063-40AC-B81A-CC0A630FCFCC",
  "TCGA-25-1627-01Z-00-DX1.7596F7F5-D705-46B4-B3A6-833D31482D28",
  "TCGA-25-1630-01Z-00-DX1.75C53351-1511-4CCE-97AC-1650010AC568",
  "TCGA-25-1634-01Z-00-DX1.6E449C67-922B-409A-84B4-581B9B468682",
  "TCGA-25-1635-01Z-00-DX1.D21D2855-43F0-4766-934A-3CE1E3A099B9",
  "TCGA-25-1870-01Z-00-DX1.780BB62D-71C1-4B21-A2F4-61AC4B0234EE",
  "TCGA-25-1871-01Z-00-DX1.1BF3077E-910F-4992-AD46-B6869A136E98",
  "TCGA-25-1878-01Z-00-DX1.618F8128-41C4-411B-9B26-B490E066B30D",
  "TCGA-25-2042-01Z-00-DX1.79198B62-0E0B-46C4-8D35-FE20582B8035",
  "TCGA-25-2390-01Z-00-DX1.BC83304E-747C-47EC-BD80-BA0B322C8630",
  "TCGA-25-2391-01Z-00-DX1.3930615C-785B-48D4-BD69-FD87932A518A",
  "TCGA-25-2392-01Z-00-DX1.C37932E5-973F-444D-8CEB-1BED4279165E",
  "TCGA-25-2393-01Z-00-DX1.A311D005-D661-4851-A54F-EE3310C9CD8A",
  "TCGA-25-2396-01Z-00-DX1.3E755B49-EFC5-49B8-A381-0B06DAE56961",
  "TCGA-25-2397-01Z-00-DX1.B6FDE9A3-6A51-4A79-82CB-51ED37945A54",
  "TCGA-25-2398-01Z-00-DX1.3F0E9A15-DBC8-43A0-8C94-20B02757CD10",
  "TCGA-25-2399-01Z-00-DX1.B9A34CB0-9EBB-4435-8EDE-E4B9078565B1",
  "TCGA-25-2400-01Z-00-DX1.AAD20514-6FF7-47A5-89DB-8AA15DCC101B",
  "TCGA-25-2401-01Z-00-DX1.AEA99982-DB6B-4178-BEDC-6E1F605B8D7C",
  "TCGA-25-2404-01Z-00-DX1.7193B185-31C3-41D0-9194-86DEDF15F096",
  "TCGA-25-2409-01Z-00-DX1.34FFD6A9-4919-4DF5-8415-215CC833F12D",
  "TCGA-3P-A9WA-01Z-00-DX1.A5FDC436-7F0D-4DD2-8644-CEF1BFF9C614",
  "TCGA-42-2587-01Z-00-DX1.7a887a16-a8bc-403b-ab71-a8b43e5613fc",
  "TCGA-57-1582-01Z-00-DX1.BDF02DAA-E520-46BD-A479-58CAA4354B1A",
  "TCGA-57-1583-01Z-00-DX1.88ACB045-1DBD-4F0B-A6D0-53B6C208805C",
  "TCGA-57-1585-01Z-00-DX1.934E0F30-F804-4CFA-92D5-41576DA6C367",
  "TCGA-57-1586-01Z-00-DX1.5C85A9B1-9ACC-4D62-BA09-91F8BDB2A058",
  "TCGA-57-1992-01Z-00-DX1.1022B1ED-3DC1-4F71-8635-0F8B28138B8C",
  "TCGA-57-1993-01Z-00-DX1.464F641B-AA9F-49A7-A224-88D9847159DF",
  "TCGA-57-1994-01Z-00-DX1.A0798185-30C5-4C32-B1A8-B440D4F4ED95",
  "TCGA-59-A5PD-01Z-00-DX1.0A4D38A2-7B15-4550-B00C-4276485A9613",
  "TCGA-5X-AA5U-01Z-00-DX1.1A03C105-4086-4AAB-B821-A9E8156FA704",
  "TCGA-OY-A56P-01Z-00-DX1.59CA76B0-5613-48CB-ACE7-6F808B8A9423",
  "TCGA-WR-A838-01Z-00-DX1.5FE22DE4-CEFB-45F6-9299-505023A8F3BA"
)

saveRDS(good_sample_names, file = "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/good_samples_name.rds")
```

# samples extraction - again but without bad samples

```{r}
ov_meta_matched <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/survival_dataset_&_metadata.rds")
good_sample_names <- readRDS("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/good_samples_name.rds")
```

```{r}
good_sample_names <- good_sample_names |> paste0(".svs")

set.seed(42) 

random_5_per_cluster <- ov_meta_matched |>
  dplyr::filter(slide_name %in% good_sample_names) |> 
  # 1. Keep only unique patients per cluster so we don't sample the same person twice
  dplyr::distinct(cluster, patientID) |>
  # 2. Group by cluster
  dplyr::group_by(cluster) |>
  # 3. Randomly pull exactly 5 rows (if a cluster has fewer than 5, it pulls all of them)
  dplyr::slice_sample(n = 5) |>
  dplyr::ungroup() |>
  # Keep just the columns you care about
  dplyr::select(cluster, patientID)

# Print the final dataframe
print(random_5_per_cluster)

# If you need them as a simple flat vector of IDs to use later:
vector_of_ids <- random_5_per_cluster$patientID
```

