---
title: "Selected samples analysis"
author: "Valerio Reffo"
date: "2026-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(anndataR)
library(ggspavis)
library(patchwork)
library(spatstat.geom)
library(spatstat)
library(sosta)
library(purrr)
library(readxl)
library(dplyr)
library(spatialFDA)
library(fda)
# survival analysis
library(survival)
library(survminer)
```

Path of the HoverNet data

```{r}
data_path <- "/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/h5ad"
img_path <- "/mnt/ganimede/shared/imageFeatureTCGA_OV/hovernet/thumb"
```

Samples to confronts are 5 patientID for each cluster defined in the samples_extraction.Rmd file
```{r}
print(vector_of_ids)
```


# Micron per Pixel extraction

Extract from xlsx file the info about mpp conversion
```{r}
xlsx_path <- "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/TCGA_data/image_metadata_TCGA.xlsx"

# 2. Read the excel file
metadata <- read_excel(xlsx_path)

# 2b. Correct row 2 because that slide is 40x, not 20x
metadata <- metadata |>
  mutate(
    # Use grepl to match the start of the image name
    `Objective Power` = if_else(grepl("TCGA-13-A5FU", `Image Name`), "40", `Objective Power`),
    AppMag            = if_else(grepl("TCGA-13-A5FU", `Image Name`), "40", AppMag),
    MPP               = if_else(grepl("TCGA-13-A5FU", `Image Name`), "0.252", MPP),
    MPP_X             = if_else(grepl("TCGA-13-A5FU", `Image Name`), "0.252", MPP_X),
    MPP_Y             = if_else(grepl("TCGA-13-A5FU", `Image Name`), "0.252", MPP_Y)
  )

# 3. Filter for your 6 samples and extract the Image Name and MPP columns
# (Note: using backticks because the column name has a space)
mpp_values <- metadata |> 
  dplyr::mutate(patientID = substr(`Image Name`, 1, 12)) |> 
  dplyr::filter(patientID %in% vector_of_ids) |> 
  dplyr::select(`Image Name`, patientID, MPP) |> 
  dplyr::mutate(MPP = as.numeric(MPP))

# 4. Convert it into a named vector so you can easily use it later
# (e.g., mpp_vector["TCGA-42-2587...svs"] will return its specific MPP)
mpp_vector <- setNames(mpp_values$MPP, mpp_values$`Image Name`)

# Print the result
print(mpp_vector)
```

## Windows

HoverNet masks reading
```{r}
mask_hovernet <- readRDS("/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/TCGA_data/window_masks_HoverNet.rds")
```

Function to transform from anndata to spatialExperiment
```{r}
ann2spe <- \(data_path, 
             spat_coords_name = c("x", "y"), 
             img_path = NULL, 
             mpp_value = NULL, ...) {
  ann <- anndataR::read_h5ad(data_path)
  #object to build SpatialExperiment
  counts <- t(ann$X) # transpose it
  cd <- ann$obs
  meta <- ann$uns
  spat <- ann$obsm$spatial
  colnames(spat) <- spat_coords_name
  cd <- cbind(cd, spat)
  spe <- SpatialExperiment::SpatialExperiment(assays = base::list("counts" = counts), 
                                              colData = cd,
                                              metadata = meta,
                                              spatialCoordsNames = spat_coords_name,
                                              imageSources = img_path, ...)
  metadata(spe)$MPP <- mpp_value
  spe
}
```

Reading all the samples togheter
```{r}
# NOT SURE IT WORK with the current_mpp (never tried)

# Use set_names() so the output list automatically uses the sample names as names for each element
spe_list <- mpp_values$`Image Name` |> 
  set_names() |> 
  map(\(sample) {
    
    # Construct the full file paths
    h5_file <- file.path(data_path, sub("\\.svs$", ".h5ad", sample))
    img_file <- file.path(img_path, sub("\\.svs$", ".png", sample)) 
    
    # Retrieve the exact MPP value for this sample from the named vector
    current_mpp <- mpp_vector[sample]
    
    # Call your custom function, now passing the mpp_value argument
    ann2spe(
      data_path = h5_file, 
      img_path = img_file,
      mpp_value = unname(current_mpp) # unname() ensures just the numeric value is passed
    )
    
  })

# Check the result
str(spe_list, max.level = 1)

#saveRDS(spe_list, "/mnt/europa/valerio/repositories/spatialPointProcess_clustering/dataset/spe_list_20slides_TCGA-OV.rds")
```

# ppp object creation 

```{r}
ppp_list <- tibble(
  name = names(spe_list),
  spe  = spe_list,
  mask = mask_hovernet[names(spe_list)] # Grabs the matching mask by name
) |> 
  # Join the metadata to get the exact AppMag for each image
  left_join(
    metadata |> select(name = `Image Name`, AppMag, MPP), 
    by = "name"
  ) |> 
  # If any slide is missing from metadata, safely default to 40
  mutate(AppMag = tidyr::replace_na(as.numeric(AppMag), 40),
         MPP    = tidyr::replace_na(as.numeric(MPP), 1)) |> 
  # Creation of ppp object 
  pmap(function(name, spe, mask, AppMag, MPP) {
  
  # Safety check
  if (is.null(mask)) {
    warning(paste("No mask found for", name))
    return(NULL)
  }
  
  # A. Extract logical matrix and transpose to [y, x]
  # (NO Y-AXIS FLIPPING - Mask stays naturally upside down to match raw cells)
  mask_mat <- mask[,,1,1] > 0.5  
  mask_spatstat <- t(mask_mat)                 
  
  # B. Create the thumbnail window
  win_thumb <- owin(
    xrange = c(0, width(mask)), 
    yrange = c(0, height(mask)), 
    mask = mask_spatstat
  )
  
  # C. Calculate exact scale directly from the injected AppMag argument
  exact_scale <- AppMag / 1.25
  
  # D. Stretch the window
  win_final <- affine(win_thumb, mat = diag(c(exact_scale, exact_scale)))
  
  # E. Build the PPP object with ORIGINAL UNTOUCHED cell coordinates
  coords <- spatialCoords(spe)
  my_ppp <- ppp(
    x = coords[, "x"], 
    y = coords[, "y"], 
    window = win_final,
    marks = as.factor(spe$type)
  )
  # F. Rescale
  if (MPP == 1) {
    warning(paste("No MPP found for", name, "- leaving in pixels"))
  }
  
  # As you correctly noted, spatstat rescale uses s = 1/mpp 
  # to multiply the underlying coordinates by mpp
  ppp_microns <- rescale(my_ppp, s = 1/MPP, unitname = c("micron", "microns"))
  
  return(my_ppp)
})

# Make sure the final list retains the sample names
names(ppp_list) <- names(spe_list)

# Check the output! The illegal points should be totally gone.
walk(ppp_list, print)
```


Plot to check if the mask is correct and the points are inside
```{r}
poinside_plot <- \(ppp){
  ggplot() +
    # 1. Plot the tissue mask (Extracts pixels on the fly)
    geom_raster(data = as.data.frame(Window(ppp)), 
                aes(x = x, y = y), fill = "grey80") +
    
    # 2. Plot the valid cells (Extracts coordinates on the fly)
    geom_point(data = as.data.frame(ppp), 
               aes(x = x, y = y), color = "blue", size = 0.1, alpha = 0.5) +
    
    # 3. Plot the rejected cells (Extracts rejects on the fly)
    geom_point(data = as.data.frame(attr(ppp, "rejects")), 
               aes(x = x, y = y), color = "red", size = 0.1, alpha = 0.5) +
    
    # Force standard Cartesian proportions so the image isn't squished
    coord_fixed() +
    theme_minimal() +
    labs(
      title = "Cell Alignment Check",
      subtitle = "Grey: Tissue Mask | Blue: Valid Cells | Red: Rejected Cells",
      x = "X Coordinate", 
      y = "Y Coordinate"
    )
}
map(ppp_list, poinside_plot)
```

# Second order statistic analysis

npoints for each cell type and potential RAM cost
```{r}
library(spatstat)
library(purrr)

# Define a safe threshold based on your RAM (e.g., 100 million pairs is usually safe for 16GB RAM)
# You might need to lower this to 50,000,000 if you have less RAM or other programs open.
MAX_PAIRS_THRESHOLD <- 100 * 10^6 

cost_npoints <- map(ppp_list, function(ppp_obj) {
  
  # 1. Extract the specific points
  n_neoplastic <- npoints(subset(ppp_obj, marks == "neoplastic"))
  n_stromal    <- npoints(subset(ppp_obj, marks == "stromal"))
  
  # 2. Calculate the "cost"
  pair_cost <- as.numeric(n_neoplastic) * as.numeric(n_stromal)
  
  # 3. Check if it exceeds the safety limit
  if (pair_cost > MAX_PAIRS_THRESHOLD) {
    message(sprintf("Skipping tile: Too large (%s pairs)", format(pair_cost, scientific = FALSE)))
    return(NULL) # Or return NA, or a placeholder
  }
  
  # 4. If safe, run the function
  tryCatch({
    Lcross(ppp_obj, i = "neoplastic", j = "stromal", correction = "best")
  }, error = function(e) {
    message("Error in Lcross: ", e$message)
    return(NULL)
  })
})
```

# memory check

Check if Lcross will crash - all unsupervised copilot code 
```{r}
library(spatstat)
library(scales) # For readable formatting (e.g., "1.2 GB")

estimate_lcross_memory <- function(ppp_obj, type_i, type_j, available_ram_gb = 60) {
  
  # 1. Check if marks exist and contain the requested types
  if (is.null(marks(ppp_obj))) {
    warning("Object has no marks.")
    return(NULL)
  }
  
  # 2. Count points
  # We use sum(marks == ...) which is faster than subsetting the whole object
  n_i <- sum(marks(ppp_obj) == type_i)
  n_j <- sum(marks(ppp_obj) == type_j)
  
  # 3. Calculate total pairs
  # We use as.numeric to prevent integer overflow for very large numbers
  n_pairs <- as.numeric(n_i) * as.numeric(n_j)
  
  # 4. Estimate Memory Usage
  # Base R double precision = 8 bytes per number.
  # Spatstat's Lcross/Kcross often needs to store:
  # - The distances (8 bytes)
  # - Edge correction weights (8 bytes)
  # - Temporary sorting/index vectors (4-8 bytes)
  # - Overhead for function calls
  # We use a safety multiplier of ~32 bytes per pair to be conservative.
  bytes_per_pair <- 32 
  est_bytes <- n_pairs * bytes_per_pair
  est_gb <- est_bytes / (1024^3)
  
  # 5. Determine if it's safe for your server
  # We leave 10% buffer for the OS and R overhead
  safe_limit <- available_ram_gb * 0.90 
  is_safe <- est_gb < safe_limit
  
  # 6. Return a summary list
  return(list(
    n_i = n_i,
    n_j = n_j,
    n_pairs = n_pairs,
    est_memory_gb = round(est_gb, 4),
    safe_to_run = is_safe,
    message = sprintf(
      "Tile Analysis: %s vs %s\n  - Count %s: %d\n  - Count %s: %d\n  - Pairs: %s\n  - Est. RAM: %s GB (Limit: %.1f GB) -> %s",
      type_i, type_j, type_i, n_i, type_j, n_j, 
      label_number(scale_cut = cut_short_scale())(n_pairs), 
      round(est_gb, 2), safe_limit,
      ifelse(is_safe, "SAFE", "DANGER")
    )
  ))
}
```

```{r}
# Example usage iterating over your list
# 1. Create a data frame of estimates
memory_check <- imap(ppp_list, function(ppp, name) {
  
  est <- estimate_lcross_memory(ppp, 
                                type_i = "neoplastic", 
                                type_j = "stromal", 
                                available_ram_gb = 60)
  
  # Return a tidy row for a dataframe
  data.frame(
    Sample = name,
    N_Neoplastic = est$n_i,
    N_Stromal = est$n_j,
    Pairs = est$n_pairs,
    Est_RAM_GB = est$est_memory_gb,
    Safe = est$safe_to_run
  )
})

# 2. Combine into a summary table
df_memory <- do.call(rbind, memory_check)

# 3. View the dangerous tiles
print(df_memory)

# 4. Run Lcross ONLY on safe tiles
# safe_tiles <- df_memory$Sample[df_memory$Safe == TRUE]
# 
# Lcross_results <- lapply(safe_tiles, function(tile_name) {
#   message("Processing: ", tile_name)
#   Lcross(ppp_sub[[tile_name]], i = "neoplastic", j = "stromal", correction = "best")
# })
```

Check memory usage based on rmax - all unsupervised copilot code 
```{r}
check_memory_rmax <- function(ppp_obj, type_i, type_j, rmax, available_gb = 60) {
  
  # 1. Get counts
  n_i <- npoints(subset(ppp_obj, marks == type_i))
  n_j <- npoints(subset(ppp_obj, marks == type_j))
  
  # 2. Get Area of the window
  area_val <- area(Window(ppp_obj))
  
  # 3. Calculate Global Density of type J (points per unit area)
  lambda_j <- n_j / area_val
  
  # 4. Estimate neighbors per point
  # Expected number of J points within rmax of an I point (assuming homogeneity)
  # Area of search circle = pi * r^2
  search_area <- pi * rmax^2
  avg_neighbors <- lambda_j * search_area
  
  # 5. Total estimated pairs to store
  # If points are highly clustered, this might be an underestimate, so we multiply by a safety factor (e.g., 2x or 3x)
  clustering_safety_factor <- 3 
  total_pairs_est <- n_i * avg_neighbors * clustering_safety_factor
  
  # 6. Convert to Memory (GB)
  # 32 bytes per pair is a conservative estimate for R vectors + overhead
  bytes_per_pair <- 32 
  mem_est_bytes <- total_pairs_est * bytes_per_pair
  mem_est_gb <- mem_est_bytes / 10^9
  
  # 7. Output results
  cat(sprintf("--- Memory Check for rmax = %s ---\n", rmax))
  cat(sprintf("Tile: %s x %s\n", type_i, type_j))
  cat(sprintf("Points: %s (i) vs %s (j)\n", n_i, n_j))
  cat(sprintf("Search Area (pi*r^2): %.2f\n", search_area))
  cat(sprintf("Est. Neighbors per point: %.2f (Global Avg)\n", avg_neighbors))
  cat(sprintf("Est. Total Pairs (with %dx cluster safety): %s\n", clustering_safety_factor, format(round(total_pairs_est), big.mark=",")))
  cat(sprintf("Est. RAM Usage: %.4f GB\n", mem_est_gb))
  
  if(mem_est_gb > (available_gb * 0.8)) {
    cat("⚠️  WARNING: MIGHT CRASH SERVER\n")
    return(FALSE)
  } else {
    cat("✅  SAFE TO RUN\n")
    return(TRUE)
  }
}

# --- Usage Example ---
map(ppp_list, \(ppp) {check_memory_rmax(ppp, "neoplastic", "stromal", rmax = 100)})
```


# Choice of rmax parameter
```{r}
rmax_list <- map(spe_list, \(spe){
  p <- rMaxHeuristic(spe, subsetby = "sample_id", marks = "type")
  p
})
walk(rmax_list, print)
```

# Reading saved data

```{r}
ppp_list <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/ppp_list20slides_TCGA-OV.rds")
names(ppp_list)
```



# Homogeneus ppp

## Lfunction estimate

L function between neoplastic and stromal
```{r}
# Lcross_homo <- map(ppp_list, \(ppp_obj) {
#      # Calculate Lcross
#      L <- Lcross(ppp_obj, i = "neoplastic", j = "stromal", correction = "border", rmax = 50)
#      L 
# }) 
```

Downsample to avoid crash

Function to estimate Lcross function for homogeneus point pattern between 2 marks and with downsample
```{r}
Lcross_personal <- function(ppp_obj, mark_i, mark_j, downsample = TRUE, n_down = 10000, ...) {
  
  # 1. Get the indices of the specific cell types dynamically
  idx_i <- which(marks(ppp_obj) == mark_i)
  idx_j <- which(marks(ppp_obj) == mark_j)
  
  # Safety check: ensure both cell types actually exist in the image!
  if(length(idx_i) < 2 | length(idx_j) < 2) {
    warning(paste("Not enough points for", mark_i, "and/or", mark_j, "to compute Lcross. Returning NULL."))
    return(NULL)
  }
  
  # 2. Handle Downsampling
  if (downsample) {
    # Randomly sample a maximum of n_down cells of each type
    keep_i <- sample(idx_i, min(length(idx_i), n_down))
    keep_j <- sample(idx_j, min(length(idx_j), n_down))
    
    # Subset the ppp object
    ppp_sub <- ppp_obj[c(keep_i, keep_j)]
    
  } else {
    # Even if we don't downsample, subsetting to JUST the two marks of interest 
    # saves a massive amount of memory for spatstat calculation
    ppp_sub <- ppp_obj[c(idx_i, idx_j)]
  }
  
  # 3. Calculate Lcross dynamically using the provided mark names
  L <- Lcross(ppp_sub, 
              i = mark_i, 
              j = mark_j, ...)
  
  return(L)
}
```


```{r}

r_values <- seq(0, 100, length.out = 40)

Lcross_benign <- map(ppp_list, \(ppp_obj) {
  Lcross_personal(ppp_obj, mark_i = "neoplastic", mark_j = "benign epithelial", 
                  downsample = TRUE, n_down = 50000, rmax = 100, correction = "border", r = r_values)
})

Lcross_stroma <- map(ppp_list, \(ppp_obj) {
  Lcross_personal(ppp_obj, mark_i = "neoplastic", mark_j = "stromal", 
                  downsample = TRUE, n_down = 50000, rmax = 100, correction = "border", r = r_values)
})
```



Plot the results

```{r}
plot_fv_list <- function(list_of_funcs, comparison_df, correction = "border") {
  
  # 1. Run the map_dfr block dynamically on whatever list you provide
  df_lcross <- map_dfr(names(list_of_funcs), function(slide_name) {
    
    # Extract the specific L-function object from the provided list
    L_obj <- list_of_funcs[[slide_name]]
    
    as.data.frame(L_obj) |> 
      mutate(
        Slide = slide_name,
        # Use .data[[correction]] to dynamically grab the 'border' column (or 'iso', etc.)
        L_r_minus_r = .data[[correction]] - r,  
        
        # Extract the marks names exactly as you designed
        mark_i = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][1]),
        mark_j = trimws(strsplit(gsub("list\\(|\\)", "", attr(L_obj, "fname")[2]), ",")[[1]][2])
      )
  })
  
  # 2. Join the metadata (comparison_benign) into the dataframe
  df_lcross <- df_lcross |> 
    mutate(
      # Extract the first 12 characters to match "TCGA-XX-XXXX"
      patientID = substr(Slide, 1, 12)
    ) |> 
    # Merge the cluster info based on the patientID
    left_join(comparison_df, by = "patientID") |> 
    # Convert cluster_ilaria to a factor so ggplot uses discrete colors instead of a continuous gradient
    mutate(cluster_ilaria = as.factor(cluster_ilaria))
  
  # 3. Build the ggplot
  # Note: group = Slide ensures each patient gets their own line
  #       color = cluster_ilaria applies the cluster colors
  p <- ggplot(df_lcross, aes(x = r, y = L_r_minus_r, group = Slide, color = cluster_ilaria)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(alpha = 0.7, linewidth = 0.6) + 
    theme_minimal() +
    labs(
      title = paste0("L-cross Function: ", df_lcross$mark_i[1], " vs ", df_lcross$mark_j[1]),
      subtitle = "Centered L-function (L(r) - r). Positive = Attraction, Negative = Repulsion",
      x = "Radius (r)",
      y = "L(r) - r",
      color = "Cluster" # Renamed legend title
    ) +
    theme(legend.position = "right")
  
  return(p)
}
```


```{r}
plot_colored <- plot_fv_list(
  list_of_funcs = Lcross_benign, 
  comparison_df = comparison_benign, 
  correction = "border"
)
plot_colored2 <- plot_fv_list(
  list_of_funcs = Lcross_stroma, 
  comparison_df = comparison_stroma, 
  correction = "border"
)

plot_colored
plot_colored2
```


Plot of the lcross functions for neoplastic vs benign epitelial and stromal
```{r}
(plcross_nb <- plot_fv_list(Lcross_benign))
(plcross_ns <- plot_fv_list(Lcross_stroma))
```

# Functional Data Analysis


# generic function for smoothing funs
```{r}
fsmooth <- function(fcross_list, fun, M = 6, genLfun.fd = 4, centrata = F, ...)
{

  # 1. Calculate raw L-functions (Common for both)
  r <- fcross_list[[1]]$r
  mat_fun <- sapply(fcross_list, function(x) x$border)
  
  # 2. Apply centering ONLY if requested
  if(centrata == TRUE){
    mat_fun <- sweep(mat_fun, MARGIN = 1, STATS = r, FUN = "-")
  }

  # 3. Create Basis and Smooth
  K <- length(r) + M - 2
  genbasis <- create.bspline.basis(range(r), K, M, r)
  genfdPar <- fdPar(genbasis, genLfun.fd, lambda = 1e-11)
  
  genfdSmooth <- smooth.basis(r, mat_fun, genfdPar)
  Genfun.fd <- genfdSmooth$fd
  
  # 4. Set Names
  fdn_dist <- paste("Distanza (da 0 a", max(r), "micron)")
  fdnames <- list(fdn_dist,
                  "Gene" = colnames(mat_fun), # Crucial: use names of surviving columns
                  "Funzione")
  Genfun.fd$fdnames <- fdnames
  
  return(list(fd = Genfun.fd, fdPar = genfdPar, 
              basis = genbasis, mat = mat_fun, r = r))
}
```



```{r}
Lfun_benign <- fsmooth(Lcross_benign, centrata = T)
Lfun_stroma <- fsmooth(Lcross_stroma, centrata = T)
```

```{r}
plot(Lfun_benign$fd)
lines(c(0,350), c(0,0), col = "gold", lwd = 2)

plot(Lfun_stroma$fd)
lines(c(0,350), c(0,0), col = "gold", lwd = 2)
```

## Comparison between neoplastic and benign.epithelial

### Functional PCA

Calcoliamo fpca per 4 componenti e mostriamo il risultato sulla rotazione VARIMAX

```{r}
fpca <- pca.fd(Lfun_benign$fd, nharm = 4, Lfun_benign$fdPar)
rownames(fpca$scores) <- names(Lcross_benign) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca)
```

### fPCA plot ruotate per VARIMAX

The plots below show the mean function and the effects of adding (+) and subtracting (−) a
suitable multiple of the first (resp. second) principal component function.

```{r}
plot.pca.fd(fpcaVarmx, cex.main=0.9)
```

## Clustering

Applichiamo due metodi di clustering sugli scores delle prime 4 componenti.
Clustering gerarchico con metodo Ward e kmeans

```{r}
set.seed(123)
fpca_km <- kmeans(fpca$scores[,c(1,4)], 4)
fpca_hc <- hclust(dist(fpca$scores[,c(1,4)]), method = "ward.D2")
memb <- cutree(fpca_hc, k = 4)
```

Grafico del dendogramma per il clustering gerarchico, scegliamo di dividere in 5 cluster.

```{r}
dhc <- as.dendrogram(fpca_hc)
plot(dhc, cex = 0.1, horiz = T)
#ggdendrogram(dhc, rotate = TRUE, size = 2)
table(kmean = fpca_km$cluster)
table(ward.d2 = memb)
```

```{r}
plot(fpca$scores[,1], fpca$scores[,2], col = fpca_km$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca$scores[,1], fpca$scores[,2], col = memb, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

```{r}
df_A <- data.frame(
  patientID = c(
    "TCGA-OY-A56P", "TCGA-23-1021", "TCGA-23-2072", "TCGA-23-1031", "TCGA-23-1117", # Cluster 1
    "TCGA-25-2401", "TCGA-25-2399", "TCGA-25-1625", "TCGA-25-1328", "TCGA-25-1871", # Cluster 2
    "TCGA-25-1315", "TCGA-25-1326", "TCGA-25-2391", "TCGA-25-1635", "TCGA-25-1627", # Cluster 3
    "TCGA-57-1582", "TCGA-13-A5FU", "TCGA-57-1993", "TCGA-57-1994", "TCGA-57-1583"  # Cluster 4
  ),
  cluster_ilaria = c(
    rep(1, 5), 
    rep(2, 5), 
    rep(3, 5), 
    rep(4, 5)
  )
)

# 3. Clean Dataset B to match IDs
# We extract the first 12 characters (e.g., "TCGA-25-1326")
df_clust <- data.frame(
  patientID = substr(names(memb), 1, 12),
  cluster_kmean = fpca_km$cluster,
  cluster_ward = memb
) |> print()

# 4. Merge to find correspondence
comparison_benign <- inner_join(df_A, df_clust, by = "patientID") |> 
  arrange(cluster_kmean, patientID)

# 5. Display the Result
print(comparison_benign)

table("ilaria_clust" = comparison_benign$cluster_ilaria, "valerio_clust" = comparison_benign$cluster_kmean)
```

## Comparison between neoplastic and stromal 

### Functional PCA

Calcoliamo fpca per 4 componenti e mostriamo il risultato sulla rotazione VARIMAX

```{r}
fpca <- pca.fd(Lfun_stroma$fd, nharm = 4, Lfun_stroma$fdPar)
rownames(fpca$scores) <- names(Lcross_stroma) # "nome" dei pazienti (numero del paziente)

fpcaVarmx <- varmx.pca.fd(fpca)
```

### fPCA plot ruotate per VARIMAX

The plots below show the mean function and the effects of adding (+) and subtracting (−) a
suitable multiple of the first (resp. second) principal component function.

```{r}
plot.pca.fd(fpcaVarmx, cex.main=0.9)
```

## Clustering

Applichiamo due metodi di clustering sugli scores delle prime 4 componenti.
Clustering gerarchico con metodo Ward e kmeans

```{r}
set.seed(123)
fpca_km <- kmeans(fpca$scores[,c(1,4)], 4)
fpca_hc <- hclust(dist(fpca$scores[,c(1,4)]), method = "ward.D2")
memb <- cutree(fpca_hc, k = 4)
```

Grafico del dendogramma per il clustering gerarchico, scegliamo di dividere in 5 cluster.

```{r}
dhc <- as.dendrogram(fpca_hc)
plot(dhc, cex = 0.1, horiz = T)
#ggdendrogram(dhc, rotate = TRUE, size = 2)
table(kmean = fpca_km$cluster)
table(ward.d2 = memb)
```

```{r}
plot(fpca$scores[,1], fpca$scores[,2], col = fpca_km$cluster, pch = 3, main = "kmeans")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
plot(fpca$scores[,1], fpca$scores[,2], col = memb, pch = 3, main = "Ward.d2")
abline(v = 0, lty = "dashed", col = "grey")
abline(h = 0, lty = "dashed", col = "grey")
```

```{r}
df_A <- data.frame(
  patientID = c(
    "TCGA-OY-A56P", "TCGA-23-1021", "TCGA-23-2072", "TCGA-23-1031", "TCGA-23-1117", # Cluster 1
    "TCGA-25-2401", "TCGA-25-2399", "TCGA-25-1625", "TCGA-25-1328", "TCGA-25-1871", # Cluster 2
    "TCGA-25-1315", "TCGA-25-1326", "TCGA-25-2391", "TCGA-25-1635", "TCGA-25-1627", # Cluster 3
    "TCGA-57-1582", "TCGA-13-A5FU", "TCGA-57-1993", "TCGA-57-1994", "TCGA-57-1583"  # Cluster 4
  ),
  cluster_ilaria = c(
    rep(1, 5), 
    rep(2, 5), 
    rep(3, 5), 
    rep(4, 5)
  )
)

# 3. Clean Dataset B to match IDs
# We extract the first 12 characters (e.g., "TCGA-25-1326")
df_clust <- data.frame(
  patientID = substr(names(vec_kmean), 1, 12),
  cluster_kmean = fpca_km$cluster,
  cluster_ward = memb
)

# 4. Merge to find correspondence
comparison_stroma <- inner_join(df_A, df_clust, by = "patientID") |> 
  arrange(cluster_kmean, patientID)

# 5. Display the Result
#print(comparison_stroma)

table("ilaria_clust" = comparison_stroma$cluster_ilaria, "valerio_clust" = comparison_stroma$cluster_kmean)
```

Comparison between the 2 results of my work
```{r}
table("benign_clust" = comparison_benign$cluster_kmean, "stroma_cluster" = comparison_stroma$cluster_kmean)
```

# Survival analysis

```{r}
ov_meta <- readRDS("~/repositories/spatialPointProcess_clustering/dataset/TCGA_data/survival_dataset_&_metadata.rds")
```


```{r}
df_combined <- ov_meta |> 
  # Select only the columns we want to add from the first df
  inner_join(df_clust |> dplyr::select(patientID, cluster_kmean), by = "patientID")
```

```{r}
fit <- survfit(Surv(time, vital_status) ~ cluster, data = df_combined)
(km_plot <- ggsurvplot(fit, data = df_combined, risk.table = TRUE, pval = TRUE, xlim = c(0,3000)))
pairwise_survdiff(Surv(time, vital_status) ~ cluster, data = df_combined)
```




COSE DA FARE:

dopo aver rimosso casi strani, prendo tutti i pazienti e faccio tutta la pipe
con i gruppi di clustering faccio survival. 

sketching: subsampling 50k cells, ribilianciare i punti con pesi

unisco benign e stromal
match i neoplastic con le altre.